---
title: "Report 12"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=4, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(stringr)
library(lubridate)
library(patchwork)
library(mosaic)
library(tidyr)
library(formatR)

library(data.table)
library(tidycensus)
library(dbarts)
library(iml)

library(caret)
library(purrr)
library(fastDummies)
library(GGally)
library(mgcv) 

library(xgboost)
```

# Introduction
This week, I will be looking at a report summarizing all the models over the first half of the semester and their appropriate metrics and results. 

# Data preprocessing
```{r}
df_complete = read.csv("../complete_metrics.csv")
df_complete = drop_na(df_complete)

set.seed(123)
n = nrow(df_complete)
train_idx = sample(seq_len(n), size = floor(0.7*n)) # 80% train
test_idx = setdiff(seq_len(n), train_idx)

vars = c("pop_ratio", "pct_urban", "density_ratio", "n_hsa", "area_ratio", "education_diff", "income_diff", "poverty_diff", "unemployment_diff")

x.train = df_complete[train_idx, vars]
y.train = df_complete$diff_wis_season[train_idx]

x.test = df_complete[test_idx, vars]
y.test = df_complete$diff_wis_season[test_idx]
```

# All horizons
```{r}
overall2 = read.csv("../forecasting_metrics_overall.csv")
pop_desc = read.csv("../us_hsa_county_popdesc.csv")
socioecon = read.csv("../Report 7/socioecon.csv") 

us_hsa_county_popdesc_bygroup = pop_desc %>%
  select(state, hsa_nci_id, area_km2_hsa, area_km2_state, n_hsa) %>% 
  distinct()

overall = overall2 %>%
  left_join(us_hsa_county_popdesc_bygroup, by = c("state", "hsa_nci_id"))

df_var_horizon = overall %>% 
  filter(season == "2023/24") %>%
  distinct()
```

```{r}
socioecon_unique = socioecon %>%
  group_by(hsa_nci_id) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

```{r}
df_horizon = left_join(df_var_horizon, socioecon_unique, by = "hsa_nci_id") %>% mutate(density_ratio = density_hsa / density_state, area_ratio = area_km2_hsa / area_km2_state)
```

```{r}
write.csv(df_horizon, "../all_horizon_complete_metrics.csv", row.names = FALSE)
```

```{r}
df_horizon = read.csv("../all_horizon_complete_metrics.csv")
df_horizon = drop_na(df_horizon)
```

Now, we have the dataset with all horizons and I think all socioecons should be included. 

## Horizon 1
### BART
```{r, fig.width = 8, fig.height = 4}
df_horizon_1 = df_horizon[df_horizon$horizon == 1,]

n = nrow(df_horizon_1)
train_idx = sample(seq_len(n), size = floor(0.8*n)) # 80% train
test_idx = setdiff(seq_len(n), train_idx)

x.train = df_horizon_1[train_idx, vars]
y.train = df_horizon_1$diff_wis_season[train_idx]

x.test = df_horizon_1[test_idx, vars]
y.test = df_horizon_1$diff_wis_season[test_idx]

bart_fit = bart(
  x.train = x.train,
  y.train = y.train,
  verbose = FALSE, keeptrees = TRUE)

pred = predict(bart_fit, newdata = x.test)
bart_predictions = as.data.frame(t(apply(pred, 2, quantile, probs = c(0.025, 0.975))))
colnames(bart_predictions) = c("ci_low", "ci_high")

bart_predictions$estimate = apply(pred, 2, mean) 
bart_predictions$observed = y.test

coverage_95_bart = mean(
  bart_predictions$observed >= bart_predictions$ci_low &
  bart_predictions$observed <= bart_predictions$ci_high,
  na.rm = TRUE
)

folds = createFolds(y.train, k = 5)

cv_R2 = c()

for (i in seq_along(folds)) {
  test_idx  = folds[[i]]
  train_idx = setdiff(seq_along(y.train), test_idx)
  
  bart_fit_cv = bart(
    x.train = x.train[train_idx, ],
    y.train = y.train[train_idx],
    verbose = FALSE,
    keeptrees = TRUE
  )
  
  pred_cv = colMeans(predict(bart_fit_cv, newdata = x.train[test_idx, ]))
  y_true  = y.train[test_idx]
  
  R2_fold = 1 - sum((y_true - pred_cv)^2) / sum((y_true - mean(y_true))^2)
  cv_R2[i] = R2_fold
}

cv_R2_bart = mean(cv_R2)

R2_bart = 1 - sum((bart_predictions$observed - bart_predictions$estimate)^2) /
  sum((bart_predictions$observed - mean(bart_predictions$observed))^2)

rmse_bart = sqrt(mean((y.test - colMeans(pred))^2))

horizon_1_bart = c("BART", R2_bart, cv_R2_bart, rmse_bart, coverage_95_bart)
```

```{r, fig.width = 8, fig.height = 4}
bart_1 = ggplot(bart_predictions, aes(x = estimate, y = observed)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, alpha = 0.35) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted (season, socioeconomic, 1)", 
       subtitle = sprintf("95%% coverage = %.1f%% (n=%d)", 100*coverage_95_bart, nrow(bart_predictions))) +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))
```

### XGBoost
```{r}
dtrain = xgb.DMatrix(data = as.matrix(x.train), label = y.train)
dtest  = xgb.DMatrix(data = as.matrix(x.test), label = y.test)

params = list(
  objective = "reg:squarederror",
  eval_metric = "rmse",
  eta = 0.05,              # lower learning rate for smoother fits
  max_depth = 2,           # shallow trees for low variance
  subsample = 0.6,         # randomly drop rows
  colsample_bytree = 0.6,  # randomly drop columns
  min_child_weight = 5,    # require decent sample size per leaf
  lambda = 5,              # L2 regularization
  alpha = 2                # L1 regularization
)

cv = xgb.cv(
  params = params,
  data = dtrain,
  nrounds = 500,
  nfold = 10,
  early_stopping_rounds = 20,
  verbose = 0
)

best_nrounds = cv$best_iteration

xgb_fit = xgb.train(
  params = params,
  data = dtrain,
  nrounds = best_nrounds,
  verbose = 0
)

pred_mean = predict(xgb_fit, as.matrix(x.test))

n = length(y.test)
p = ncol(x.train)
R2 = 1 - sum((y.test - pred_mean)^2) / sum((y.test - mean(y.test))^2)

RMSE = sqrt(mean((y.test - pred_mean)^2))

cv_mean_R2 = 1 - (cv$evaluation_log$test_rmse_mean[cv$best_iteration]^2 /
                  var(y.train))

metrics = data.frame (
  Metric = c("$R^2$", "Cross-validated $R^2$", "RMSE"),
  Value = round(c(R2, cv_mean_R2, RMSE),4)
)

xgb_predictions = data.frame(
  estimate = pred_mean,
  observed = y.test
)

horizon_1_xgb = c("XGBoost", "N/A", cv_mean_R2, RMSE, "N/A")
```

```{r, fig.width = 8, fig.height = 4}
xgb_1 = ggplot(xgb_predictions, aes(x = estimate, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Predicted diff_wis_season",
    y = "Observed diff_wis_season",
    title = "Observed vs Predicted (season, XGBoost, 1)",
    subtitle = sprintf("R-squared = %.2f | CV R-squared = %.2f | RMSE = %.2f | n = %.0f", R2, cv_mean_R2, RMSE, nrow(x.test))
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(-1, 2, by = 0.25)) +
  coord_cartesian(xlim = c(-0.25, 1))
```

## Horizon 2
### BART
```{r, fig.width = 8, fig.height = 4}
df_horizon_2 = df_horizon[df_horizon$horizon == 2,]

n = nrow(df_horizon_2)
train_idx = sample(seq_len(n), size = floor(0.8*n)) # 80% train
test_idx = setdiff(seq_len(n), train_idx)

x.train = df_horizon_2[train_idx, vars]
y.train = df_horizon_2$diff_wis_season[train_idx]

x.test = df_horizon_2[test_idx, vars]
y.test = df_horizon_2$diff_wis_season[test_idx]

bart_fit = bart(
  x.train = x.train,
  y.train = y.train,
  verbose = FALSE, keeptrees = TRUE)

pred = predict(bart_fit, newdata = x.test)
bart_predictions = as.data.frame(t(apply(pred, 2, quantile, probs = c(0.025, 0.975))))
colnames(bart_predictions) = c("ci_low", "ci_high")

bart_predictions$estimate = apply(pred, 2, mean) 
bart_predictions$observed = y.test

coverage_95_bart = mean(
  bart_predictions$observed >= bart_predictions$ci_low &
  bart_predictions$observed <= bart_predictions$ci_high,
  na.rm = TRUE
)

folds = createFolds(y.train, k = 5)

cv_R2 = c()

for (i in seq_along(folds)) {
  test_idx  = folds[[i]]
  train_idx = setdiff(seq_along(y.train), test_idx)
  
  bart_fit_cv = bart(
    x.train = x.train[train_idx, ],
    y.train = y.train[train_idx],
    verbose = FALSE,
    keeptrees = TRUE
  )
  
  pred_cv = colMeans(predict(bart_fit_cv, newdata = x.train[test_idx, ]))
  y_true  = y.train[test_idx]
  
  R2_fold = 1 - sum((y_true - pred_cv)^2) / sum((y_true - mean(y_true))^2)
  cv_R2[i] = R2_fold
}

cv_R2_bart = mean(cv_R2)

R2_bart = 1 - sum((bart_predictions$observed - bart_predictions$estimate)^2) /
  sum((bart_predictions$observed - mean(bart_predictions$observed))^2)

rmse_bart = sqrt(mean((y.test - colMeans(pred))^2))

horizon_2_bart = c("BART", R2_bart, cv_R2_bart, rmse_bart, coverage_95_bart)
```

```{r, fig.width = 8, fig.height = 4}
bart_2 = ggplot(bart_predictions, aes(x = estimate, y = observed)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, alpha = 0.35) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted (season, socioeconomic, 2)",
       subtitle = sprintf("95%% coverage = %.1f%% (n=%d)", 100*coverage_95_bart, nrow(bart_predictions))) +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))
```

### XGBoost
```{r}
dtrain = xgb.DMatrix(data = as.matrix(x.train), label = y.train)
dtest  = xgb.DMatrix(data = as.matrix(x.test), label = y.test)

params = list(
  objective = "reg:squarederror",
  eval_metric = "rmse",
  eta = 0.05,              # lower learning rate for smoother fits
  max_depth = 2,           # shallow trees for low variance
  subsample = 0.6,         # randomly drop rows
  colsample_bytree = 0.6,  # randomly drop columns
  min_child_weight = 5,    # require decent sample size per leaf
  lambda = 5,              # L2 regularization
  alpha = 2                # L1 regularization
)

cv = xgb.cv(
  params = params,
  data = dtrain,
  nrounds = 500,
  nfold = 10,
  early_stopping_rounds = 20,
  verbose = 0
)

best_nrounds = cv$best_iteration

xgb_fit = xgb.train(
  params = params,
  data = dtrain,
  nrounds = best_nrounds,
  verbose = 0
)

pred_mean = predict(xgb_fit, as.matrix(x.test))

n = length(y.test)
p = ncol(x.train)
R2 = 1 - sum((y.test - pred_mean)^2) / sum((y.test - mean(y.test))^2)

RMSE = sqrt(mean((y.test - pred_mean)^2))

cv_mean_R2 = 1 - (cv$evaluation_log$test_rmse_mean[cv$best_iteration]^2 /
                  var(y.train))

metrics = data.frame (
  Metric = c("$R^2$", "Cross-validated $R^2$", "RMSE"),
  Value = round(c(R2, cv_mean_R2, RMSE),4)
)

xgb_predictions = data.frame(
  estimate = pred_mean,
  observed = y.test
)

horizon_2_xgb = c("XGBoost", "N/A", cv_mean_R2, RMSE, "N/A")
```

```{r, fig.width = 8, fig.height = 4}
xgb_2 = ggplot(xgb_predictions, aes(x = estimate, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Predicted diff_wis_season",
    y = "Observed diff_wis_season",
    title = "Observed vs Predicted (season, XGBoost, 2)",
    subtitle = sprintf("R-squared = %.2f | CV R-squared = %.2f | RMSE = %.2f | n = %.0f", R2, cv_mean_R2, RMSE, nrow(x.test))
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(-1, 2, by = 0.25)) +
  coord_cartesian(xlim = c(-0.25, 1))
```

## Horizon 3
### BART
```{r, fig.width = 8, fig.height = 4}
df_horizon_3 = df_horizon[df_horizon$horizon == 3,]

n = nrow(df_horizon_3)
train_idx = sample(seq_len(n), size = floor(0.8*n)) # 80% train
test_idx = setdiff(seq_len(n), train_idx)

x.train = df_horizon_3[train_idx, vars]
y.train = df_horizon_3$diff_wis_season[train_idx]

x.test = df_horizon_3[test_idx, vars]
y.test = df_horizon_3$diff_wis_season[test_idx]

bart_fit = bart(
  x.train = x.train,
  y.train = y.train,
  verbose = FALSE, keeptrees = TRUE)

pred = predict(bart_fit, newdata = x.test)
bart_predictions = as.data.frame(t(apply(pred, 2, quantile, probs = c(0.025, 0.975))))
colnames(bart_predictions) = c("ci_low", "ci_high")

bart_predictions$estimate = apply(pred, 2, mean) 
bart_predictions$observed = y.test

coverage_95_bart = mean(
  bart_predictions$observed >= bart_predictions$ci_low &
  bart_predictions$observed <= bart_predictions$ci_high,
  na.rm = TRUE
)

folds = createFolds(y.train, k = 5)

cv_R2 = c()

for (i in seq_along(folds)) {
  test_idx  = folds[[i]]
  train_idx = setdiff(seq_along(y.train), test_idx)
  
  bart_fit_cv = bart(
    x.train = x.train[train_idx, ],
    y.train = y.train[train_idx],
    verbose = FALSE,
    keeptrees = TRUE
  )
  
  pred_cv = colMeans(predict(bart_fit_cv, newdata = x.train[test_idx, ]))
  y_true  = y.train[test_idx]
  
  R2_fold = 1 - sum((y_true - pred_cv)^2) / sum((y_true - mean(y_true))^2)
  cv_R2[i] = R2_fold
}

cv_R2_bart = mean(cv_R2)

R2_bart = 1 - sum((bart_predictions$observed - bart_predictions$estimate)^2) /
  sum((bart_predictions$observed - mean(bart_predictions$observed))^2)

rmse_bart = sqrt(mean((y.test - colMeans(pred))^2))

horizon_3_bart = c("BART", R2_bart, cv_R2_bart, rmse_bart, coverage_95_bart)
```

```{r, fig.width = 8, fig.height = 4}
bart_3 = ggplot(bart_predictions, aes(x = estimate, y = observed)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, alpha = 0.35) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted (season, socioeconomic, 3)",
       subtitle = sprintf("95%% coverage = %.1f%% (n=%d)", 100*coverage_95_bart, nrow(bart_predictions))) +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))
```

### XGBoost
```{r}
dtrain = xgb.DMatrix(data = as.matrix(x.train), label = y.train)
dtest  = xgb.DMatrix(data = as.matrix(x.test), label = y.test)

params = list(
  objective = "reg:squarederror",
  eval_metric = "rmse",
  eta = 0.05,              # lower learning rate for smoother fits
  max_depth = 2,           # shallow trees for low variance
  subsample = 0.6,         # randomly drop rows
  colsample_bytree = 0.6,  # randomly drop columns
  min_child_weight = 5,    # require decent sample size per leaf
  lambda = 5,              # L2 regularization
  alpha = 2                # L1 regularization
)

cv = xgb.cv(
  params = params,
  data = dtrain,
  nrounds = 500,
  nfold = 10,
  early_stopping_rounds = 20,
  verbose = 0
)

best_nrounds = cv$best_iteration

xgb_fit = xgb.train(
  params = params,
  data = dtrain,
  nrounds = best_nrounds,
  verbose = 0
)

pred_mean = predict(xgb_fit, as.matrix(x.test))

n = length(y.test)
p = ncol(x.train)
R2 = 1 - sum((y.test - pred_mean)^2) / sum((y.test - mean(y.test))^2)

RMSE = sqrt(mean((y.test - pred_mean)^2))

cv_mean_R2 = 1 - (cv$evaluation_log$test_rmse_mean[cv$best_iteration]^2 /
                  var(y.train))

metrics = data.frame (
  Metric = c("$R^2$", "Cross-validated $R^2$", "RMSE"),
  Value = round(c(R2, cv_mean_R2, RMSE),4)
)

xgb_predictions = data.frame(
  estimate = pred_mean,
  observed = y.test
)

horizon_3_xgb = c("XGBoost", "N/A", cv_mean_R2, RMSE, "N/A")
```

```{r, fig.width = 8, fig.height = 4}
xgb_3 = ggplot(xgb_predictions, aes(x = estimate, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Predicted diff_wis_season",
    y = "Observed diff_wis_season",
    title = "Observed vs Predicted (season, XGBoost, 3)",
    subtitle = sprintf("R-squared = %.2f | CV R-squared = %.2f | RMSE = %.2f | n = %.0f", R2, cv_mean_R2, RMSE, nrow(x.test))
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(-1, 2, by = 0.25)) +
  coord_cartesian(xlim = c(-0.25, 1))
```

## Horizon 4
```{r, fig.width = 8, fig.height = 4}
df_horizon_4 = df_horizon[df_horizon$horizon == 4,]

n = nrow(df_horizon_4)
train_idx = sample(seq_len(n), size = floor(0.8*n)) # 80% train
test_idx = setdiff(seq_len(n), train_idx)

x.train = df_horizon_4[train_idx, vars]
y.train = df_horizon_4$diff_wis_season[train_idx]

x.test = df_horizon_4[test_idx, vars]
y.test = df_horizon_4$diff_wis_season[test_idx]

bart_fit = bart(
  x.train = x.train,
  y.train = y.train,
  verbose = FALSE, keeptrees = TRUE)

pred = predict(bart_fit, newdata = x.test)
bart_predictions = as.data.frame(t(apply(pred, 2, quantile, probs = c(0.025, 0.975))))
colnames(bart_predictions) = c("ci_low", "ci_high")

bart_predictions$estimate = apply(pred, 2, mean) 
bart_predictions$observed = y.test

coverage_95_bart = mean(
  bart_predictions$observed >= bart_predictions$ci_low &
  bart_predictions$observed <= bart_predictions$ci_high,
  na.rm = TRUE
)

folds = createFolds(y.train, k = 5)

cv_R2 = c()

for (i in seq_along(folds)) {
  test_idx  = folds[[i]]
  train_idx = setdiff(seq_along(y.train), test_idx)
  
  bart_fit_cv = bart(
    x.train = x.train[train_idx, ],
    y.train = y.train[train_idx],
    verbose = FALSE,
    keeptrees = TRUE
  )
  
  pred_cv = colMeans(predict(bart_fit_cv, newdata = x.train[test_idx, ]))
  y_true  = y.train[test_idx]
  
  R2_fold = 1 - sum((y_true - pred_cv)^2) / sum((y_true - mean(y_true))^2)
  cv_R2[i] = R2_fold
}

cv_R2_bart = mean(cv_R2)

R2_bart = 1 - sum((bart_predictions$observed - bart_predictions$estimate)^2) /
  sum((bart_predictions$observed - mean(bart_predictions$observed))^2)

rmse_bart = sqrt(mean((y.test - colMeans(pred))^2))

horizon_4_bart = c("BART",R2_bart, cv_R2_bart, rmse_bart, coverage_95_bart)
```

### BART
```{r, fig.width = 8, fig.height = 4}
bart_4 = ggplot(bart_predictions, aes(x = estimate, y = observed)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, alpha = 0.35) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted (season, socioeconomic, 4)", 
       subtitle = sprintf("95%% coverage = %.1f%% (n=%d)", 100*coverage_95_bart, nrow(bart_predictions))) +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))
```

### XGBoost
```{r}
dtrain = xgb.DMatrix(data = as.matrix(x.train), label = y.train)
dtest  = xgb.DMatrix(data = as.matrix(x.test), label = y.test)

params = list(
  objective = "reg:squarederror",
  eval_metric = "rmse",
  eta = 0.05,              # lower learning rate for smoother fits
  max_depth = 2,           # shallow trees for low variance
  subsample = 0.6,         # randomly drop rows
  colsample_bytree = 0.6,  # randomly drop columns
  min_child_weight = 5,    # require decent sample size per leaf
  lambda = 5,              # L2 regularization
  alpha = 2                # L1 regularization
)

cv = xgb.cv(
  params = params,
  data = dtrain,
  nrounds = 500,
  nfold = 10,
  early_stopping_rounds = 20,
  verbose = 0
)

best_nrounds = cv$best_iteration

xgb_fit = xgb.train(
  params = params,
  data = dtrain,
  nrounds = best_nrounds,
  verbose = 0
)

pred_mean = predict(xgb_fit, as.matrix(x.test))

n = length(y.test)
p = ncol(x.train)
R2 = 1 - sum((y.test - pred_mean)^2) / sum((y.test - mean(y.test))^2)

RMSE = sqrt(mean((y.test - pred_mean)^2))

cv_mean_R2 = 1 - (cv$evaluation_log$test_rmse_mean[cv$best_iteration]^2 /
                  var(y.train))

metrics = data.frame (
  Metric = c("$R^2$", "Cross-validated $R^2$", "RMSE"),
  Value = round(c(R2, cv_mean_R2, RMSE),4)
)

xgb_predictions = data.frame(
  estimate = pred_mean,
  observed = y.test
)

horizon_4_xgb = c("XGBoost", "N/A", cv_mean_R2, RMSE, "N/A")
```

```{r, fig.width = 8, fig.height = 4}
xgb_4 = ggplot(xgb_predictions, aes(x = estimate, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Predicted diff_wis_season",
    y = "Observed diff_wis_season",
    title = "Observed vs Predicted (season, XGBoost, 4)",
    subtitle = sprintf("R-squared = %.2f | CV R-squared = %.2f | RMSE = %.2f | n = %.0f", R2, cv_mean_R2, RMSE, nrow(x.test))
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(-1, 2, by = 0.25)) +
  coord_cartesian(xlim = c(-0.25, 1))
```

# Results
## Table
```{r}
total_metrics = as.data.frame(rbind(horizon_1_bart, horizon_1_xgb, horizon_2_bart, horizon_2_xgb, horizon_3_bart, horizon_3_xgb,  horizon_4_bart, horizon_4_xgb), row.names = FALSE) %>% mutate(across(everything(), ~ na_if(.x, "N/A")))
total_metrics$V2 = as.numeric(total_metrics$V2)
total_metrics$V3 = as.numeric(total_metrics$V3)
total_metrics$V4 = as.numeric(total_metrics$V4)
total_metrics$V5 = as.numeric(total_metrics$V5)
total_metrics = total_metrics %>% mutate(across(where(is.numeric), ~ round(.x, 3)))
total_metrics = cbind( c("1", "1", "2", "2", "3", "3", "4", "4"), total_metrics)
colnames(total_metrics) = c("Horizon", "Model", "R2", "CVR2", "RMSE", "coverage")

knitr::kable(total_metrics, col.names = c("Horizon", "Model", "$R^2$", "CV $R^2$", "RMSE", "95% Coverage Rate"), caption = "Prediction metrics for each horizon with BART and XGBoost")
```

## Prediction graphs
```{r, fig.width =10, fig.height= 8}
(bart_1 + bart_2) / (bart_3 + bart_4)
(xgb_1 + xgb_2) / (xgb_3 + xgb_4)
```

## Metric graphs
```{r, fig.width = 8, fig.height = 8, warning = FALSE}
ggtotal = ggplot(total_metrics)

cvr2graph = ggtotal + geom_col(aes(x = reorder(paste0("H", Horizon, "-", Model), CVR2), y = CVR2, fill = Model), col = "black") + labs(title = "Cross-validated R squared by model/horizon combo", x="Model/Horizon", y= "CV R^2") + theme_minimal()

RMSEgraph = ggtotal + geom_col(aes(x = reorder(paste0("H", Horizon, "-", Model), -RMSE), y = RMSE, fill = Model), col = "black") + labs(title = "RMSE by model/horizon combo", x="Model/Horizon", y= "RMSE") + theme_minimal()

coveragegraph = ggplot(total_metrics[total_metrics$Model == "BART",]) + geom_col(aes(x = reorder(paste0("H", Horizon, "-", Model), coverage), y = coverage, fill = Model), col = "black") + labs(title = "Coverage by BART/horizon combo", x="Model/Horizon", y= "Coverage") + theme_minimal()

cvr2graph / RMSEgraph / coveragegraph
```


# Conclusion
As we can see, XGBoost has consistently higher cross-validated $R^2$ (generalizes better across horizons), BART has a weird fall-off from $R^2$ to the cross-validated version, RMSE values go down for BART with each horizon (which is curious since they get longer), and the coverage BART rates stay strong.

For feature importance (or similar feature importance) of these two models, try looking at Report 11's XGBoost section (right after the Data Preprocessing section) or Report 7's SHAP feature importance section (last section of the report).
