---
title: "Report 3"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=8, fig.width=8, fig.align = "center", warning=FALSE, echo=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r, results='hide', warning=FALSE, message=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(patchwork)
library(corrplot)
library(mosaic)
library(moderndive)
library(effectsize)
```

# Introduction
This week, I plan to use the methods I mentioned last week to estimate the WIS differences. I'll process and prepare the data and then apply several models to try and see which work best. 

# Data processing

```{r}



```

# Support Vector Regression (SVR)
## Theoretical background
## Implementation

# Bayesian Additive Regression Trees
## Theoretical background
## Implementation

# Elastic Net / LASSO / Ridge
## Theoretical background
## Implementation

# Conclusion

