
title: "Report 11"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red


```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height = 4, fig.width = 4, fig.align = "left",
                      echo = TRUE, warning = FALSE, message = FALSE,
                      tidy = TRUE, tidy.opts = list(width.cutoff = 60))
```



```{r libraries, echo=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(kableExtra)
library(sf)
```

# Introduction
This week, I was able to add some per-season maps, add some consistency plots, look at the scatter plot by urbanicity quartile, and trim the code file into a separate file for easy use.

# Data Loading

```{r load_cache}
d = readRDS("../report_data_cache.rds")
list2env(d, envir = .GlobalEnv)

lower48 = setdiff(c(state.name, "District of Columbia"), c("Alaska", "Hawaii"))

cat("Filtered rows:", nrow(diff_peak_filtered), "\n")

poster_theme =
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 20, face = "bold"),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18),
    strip.text = element_text(size = 18), # for facets
    plot.margin = margin(10, 10, 10, 40)
  )
```

# Time
```{r timing_scatter, fig.height=6, fig.width=12}
diff_peak_time = diff_peak_data |>
    filter(!is.na(centroid_time_diff) & !is.na(signed_lag))

diff_peak_time_zoom = diff_peak_time |>
    filter(signed_lag >= -10, signed_lag <= 10, centroid_time_diff >=
        -10, centroid_time_diff <= 10)

centroid_lm_zoom = lm(centroid_time_diff ~ signed_lag * factor(peak_window),
    data = diff_peak_time_zoom)

diff_peak_time_zoom$pred <- predict(centroid_lm_zoom, newdata = diff_peak_time_zoom)

timing = ggplot(diff_peak_time_zoom, aes(x = signed_lag, y = centroid_time_diff,
    fill = as.factor(peak_window))) + geom_point(pch = 21, color = "black", size = 2, stroke = 0.8) +
    geom_line(color = "black", aes(y = pred), lwd = 1.2) +
    theme_bw() + facet_wrap(~peak_window, 
        labeller = labeller(peak_window = c("1" = "2022/23", "2" = "2023/24", "3" = "2024/25"))) +
    labs(x = "DTW signed lag", y = "True timing lag (centroid)") + 
    poster_theme + theme(legend.position = "none")

ggsave("timing.png", dpi = 600, width = 12, height = 6)
```

# Magnitude

```{r mag_scatter, fig.height=6, fig.width=12}
diff_peak_mag = diff_peak_data |>
    filter(state_peak_value > 0, hsa_peak_value > 0, frechet <
        25, truth_mag_ratio < 5)  #, dtw_rmse < 5)

mag = ggplot(diff_peak_mag, aes(x = frechet, y = truth_mag_ratio)) +
    geom_point(aes(fill = as.factor(peak_window)), pch = 21, color = "black", size = 3, stroke = 0.8) + 
    geom_smooth(method = "lm", color = 'black', lwd=1.2, se = FALSE) + 
    facet_wrap(~peak_window,
        labeller = labeller(peak_window = c("1" = "2022/23", "2" = "2023/24", "3" = "2024/25"))) +
    theme_bw() + 
    labs(x = "Fréchet distance", y = "True magnitude ratio",
        fill = "Peak window", color = "Peak window") + 
    poster_theme + 
    theme(legend.position = "none")

ggsave("mag.png", dpi = 600, width = 12, height = 6)
```

# Local Texas comparisons
```{r}
hsa_state = read.csv("../hsa_state_inc.csv")
source("../metric_functions.R")
source("../graphstat_functions.R")
source("../write_dtw.R")
```

```{r, fig.width = 10, fig.height = 10}
hsa_id = 453
hsa = hsa_state[hsa_state$hsa_nci_id == hsa_id, ]

P = as.matrix(hsa$inc_hsa)
Q = as.matrix(hsa$inc_state)

res_dtw = dtw_distance(P, Q)

res = discrete_frechet(P, Q)


# dtw adjusted
P_warped = rep(NA, length(Q))
i = res_dtw$path[,1]
j = res_dtw$path[,2]
P_warped[j] = P[i]
#P_warped = na.approx(P_warped, rule = 2)

df_plot = data.frame(
  index = seq_len(length(Q)),
  HSA_aligned = P_warped,
  State = Q
)

start_date = as.Date("2022-09-26")  # Monday of 2022 Week 39
dates = start_date + (0:147) * 7    # 148 weekly dates
breaks = seq(as.Date("2022-10-01"), as.Date("2025-06-01"), by = "6 months")

normal = ggplot(hsa) +
  geom_line(linewidth = 1, aes(x = dates, y = inc_hsa, color = "HSA")) +
  geom_line(linewidth = 1, aes(x = dates, y = inc_state, color = "State")) +
  labs(
    x = "Date",
    y = "Percent influenza ED visits",
    title = "Local vs. State (Dallas)"
  ) +
  theme_minimal() +
  poster_theme +
  theme(plot.title = element_text(size = 22, face = "bold")) +
  scale_color_manual(name = "Legend", values = c(HSA = "blue", State = "red")) +
  scale_x_date(breaks = breaks, date_labels = "%b %Y")

warped = ggplot(df_plot) +
  geom_line(aes(x = dates[index], y = HSA_aligned, color = "HSA"), linewidth = 1) +
  geom_line(aes(x = dates[index], y = State, color = "State"), linewidth = 1) +
  labs(
    x = "Date",
    y = "Percent influenza ED visits",
    title = "DTW-warped Local vs. State (Dallas)"
  ) +
  theme_minimal() +
  poster_theme +
  theme(plot.title = element_text(size = 22, face = "bold")) +
  scale_color_manual(name = "Legend", values = c("HSA" = "blue", State = "red")) +
  scale_x_date(breaks = breaks, date_labels = "%b %Y")

(normal / warped)
```

```{r}
ggsave("dallas.png", dpi = 300, width = 10, height = 10)
```

# US Maps
```{r map_setup}
win_labels = c("1" = "2022/23", "2" = "2023/24", "3" = "2024/25")

# Per-window: raw metric averages (correlations would have only 1 row per group)
hsa_window_vals = diff_peak_filtered %>%
  group_by(ID, state, peak_window) %>%
  summarise(
    mean_signed_lag = mean(signed_lag,      na.rm = TRUE),
    mean_frechet    = mean(frechet,         na.rm = TRUE),
    mean_mag_ratio  = mean(truth_mag_ratio, na.rm = TRUE),
    .groups = "drop"
  )

us_map_window = us_map %>%
  filter(state %in% lower48) %>%
  crossing(peak_window = c(1, 2, 3)) %>%
  left_join(hsa_window_vals %>% rename(hsa_nci_id = ID),
            by = c("hsa_nci_id", "state", "peak_window"))

seasonal_map = function(data, fill_var, title, limits, fill_scale) {
  ggplot() +
    geom_sf(data = st_set_geometry(st_as_sf(data), "geometry_hsa"),
            aes(fill = .data[[fill_var]]), color = "gray20", linewidth = 0.1) +
    geom_sf(data = st_set_geometry(st_as_sf(data), "geometry"),
            fill = NA, color = "grey20", linewidth = 0.25) +
    fill_scale +
    facet_wrap(~peak_window, ncol = 1,
               labeller = labeller(peak_window = win_labels)) +
    coord_sf() + theme_void() +
    theme(
      strip.text = element_text(size = 20, face = "bold", margin = margin(b = 6)),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12),
      legend.key.width = unit(0.7, "cm"),
      legend.key.height = unit(0.4, "cm"),
      legend.position = "right",
      plot.margin = margin(5, 5, 5, 5)
    )
}
```

```{r map_timing, fig.height=12, fig.width=8}
seasonal_map(
  us_map_window, "mean_signed_lag",
  title = NULL,
  limits = c(-1, 1),
  fill_scale = scale_fill_gradient2(
    low = "blue", mid = "white", high = "red", midpoint = 0,
    limits = c(-1, 1), oob = scales::squish,
    na.value = "gray90", name = "Signed lag (days)"
  )
)
```

```{r}
ggsave("timingmap.png", dpi = 300, width = 8, height = 12)
```

```{r map_frechet, fig.height=12, fig.width=8}
seasonal_map(
  us_map_window, "mean_frechet",
  title = NULL,
  limits = c(0, 25),
  fill_scale = scale_fill_viridis_c(
    option = "mako", direction = -1,
    limits = c(0, 25), oob = scales::squish,
    na.value = "gray90", name = "Fréchet distance"
  )
)
```

```{r}
ggsave("frechetmap.png", dpi = 300, width = 8, height = 12)
```
