---
title: "Report 12"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=4, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(patchwork)
library(zoo)
library(kableExtra)
library(corrplot)
```

# Introduction
This week, I worked on the URF poster and made some good diagrams and then for the report, I made two animations!

# Data Loading
```{r load_cache, message = FALSE, warning = FALSE}
d = readRDS("../report_data_cache.rds")
list2env(d, envir = .GlobalEnv)
lower48 = setdiff(c(state.name, "District of Columbia"), c("Alaska", "Hawaii"))

us_map_weekly = us_map %>%
  filter(state %in% lower48) %>%
  left_join(
    weekly_map_data %>% rename(hsa_nci_id = ID),
    by = c("hsa_nci_id", "state")
  )
```

# Animation
this took way too long to save icl  
```{r}
library(gganimate)
library(gifski)

win_labels = c("1" = "2022/23", "2" = "2023/24", "3" = "2024/25")
```

# magnitude
```{r}
p_mag = ggplot() +
  geom_sf(data = st_set_geometry(st_as_sf(us_map_weekly), "geometry_hsa"),
          aes(fill = weekly_mag_ratio), color = "gray20", linewidth = 0.1) +
  geom_sf(data = st_set_geometry(st_as_sf(us_map_weekly), "geometry"),
          fill = NA, color = "grey20", linewidth = 0.25) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red", midpoint = 1,
    limits = c(0.5, 2), oob = scales::squish,
    na.value = "gray90", name = "HSA / State\nincidence"
  ) +
  facet_wrap(~peak_window, ncol = 3,
             labeller = labeller(peak_window = win_labels)) +
  coord_sf() + theme_void() +
  labs(title = "Week {closest_state}") +
  theme(
    strip.text  = element_text(size = 14, face = "bold", margin = margin(b = 6)),
    plot.title  = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 12),
    legend.text  = element_text(size = 10),
    legend.position = "right",
    plot.margin = margin(5, 5, 5, 5)
  ) +
  transition_states(week_index, transition_length = 1, state_length = 2) +
  ease_aes("linear")

animate(
  p_mag,
  nframes  = n_distinct(weekly_map_data$week_index),
  fps      = 4,
  width    = 1400, height = 500,
  renderer = gifski_renderer("weekly_mag_ratio.gif")
)
```

# lag
```{r}
p_lag = ggplot() +
  geom_sf(data = st_set_geometry(st_as_sf(us_map_weekly), "geometry_hsa"),
          aes(fill = lead_lag), color = "gray20", linewidth = 0.1) +
  geom_sf(data = st_set_geometry(st_as_sf(us_map_weekly), "geometry"),
          fill = NA, color = "grey20", linewidth = 0.25) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red", midpoint = 0,
    limits = c(-0.5, 0.5), oob = scales::squish,
    na.value = "gray90", name = "Cumulative\nlead/lag"
  ) +
  facet_wrap(~peak_window, ncol = 3,
             labeller = labeller(peak_window = win_labels)) +
  coord_sf() + theme_void() +
  labs(title = "Week {closest_state}") +
  theme(
    strip.text  = element_text(size = 14, face = "bold", margin = margin(b = 6)),
    plot.title  = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 12),
    legend.text  = element_text(size = 10),
    legend.position = "right",
    plot.margin = margin(5, 5, 5, 5)
  ) +
  transition_states(week_index, transition_length = 1, state_length = 2) +
  ease_aes("linear")
```

```{r}
animate(
  p_lag,
  nframes  = n_distinct(weekly_map_data$week_index),
  fps      = 4,
  width    = 1400, height = 500,
  renderer = gifski_renderer("weekly_lead_lag.gif")
)
```


# Conclusion
Animations look good, next week will do the choropleth of truth vs metric to see if differences are transferred / maintained from the truth to representative metrics.