---
title: "Report 8"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=4, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(stringr)
library(lubridate)
library(patchwork)
library(mosaic)
library(tidyr)
library(formatR)

library(data.table)
library(tidycensus)
library(dbarts)
library(iml)

library(caret)
library(purrr)
library(fastDummies)
library(GGally)
library(mgcv) 
```

# Introduction
This week, we are going to introduce weather data, test GLMs, test new metrics, and generate US maps. 

# Data processing
## Previous data
```{r}
df_complete = read.csv("../complete_metrics.csv")
df_complete = drop_na(df_complete)
colnames(df_complete)
```

```{r}
set.seed(123)
n = nrow(df_complete)
train_idx = sample(seq_len(n), size = floor(0.8*n)) # 80% train
test_idx = setdiff(seq_len(n), train_idx)

vars_std = c("pop_ratio", "pct_urban", "density_ratio", "n_hsa", "area_ratio")

x.train_std = df_complete[train_idx, vars_std]
y.train_std = df_complete$diff_wis_season[train_idx]

x.test_std = df_complete[test_idx, vars_std]
y.test_std = df_complete$diff_wis_season[test_idx]

vars_soc = c("pop_ratio", "pct_urban", "density_ratio", "n_hsa", "area_ratio", "education_diff", "income_diff", "poverty_diff", "unemployment_diff")

x.train_soc = df_complete[train_idx, vars_soc]
y.train_soc = df_complete$diff_wis_season[train_idx]

x.test_soc = df_complete[test_idx, vars_soc]
y.test_soc = df_complete$diff_wis_season[test_idx]
```

# Other Models 
## GLM
```{r glm, fig.width = 8, fig.height = 4}
step_glm <- glm(diff_wis_season~ pop_ratio + pct_urban + density_ratio + n_hsa + area_ratio,
      data = df_complete,
      family = gaussian(link = "identity")
    )
summary(step_glm)
#plot(step_glm)


df_fitted_glm <- data.frame(fitted = step_glm$fitted.values, observed = step_glm$y)

cor(df_fitted_glm$fitted, df_fitted_glm$observed, use = "complete.obs")^2

ggplot(df_fitted_glm, aes(x = fitted, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted (glm, season)") +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))
```

## GAM
```{r gam, fig.width = 8, fig.height = 4}
fit_final <- gam(
  diff_wis_season ~ pop_ratio + pct_urban + density_ratio + n_hsa + area_ratio,
  data = df_complete
)
summary(fit_final)

gam.check(fit_final)

df_fitted_gam <- data.frame(fitted = fit_final$fitted.values, observed = fit_final$y)

ggplot(df_fitted_gam, aes(x = fitted, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted (gam, season)") +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))

cor(df_fitted_gam$fitted, df_fitted_gam$observed, use = "complete.obs")^2

df_diag <- data.frame(
  fitted = fitted(fit_final),
  residuals = resid(fit_final, type = "deviance"),
  standardized = resid(fit_final, type = "pearson")
)

ggplot(df_diag, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Fitted values", y = "Deviance residuals")

qqnorm(df_diag$residuals)
qqline(df_diag$residuals, col = "red")
```

## BART
### Standard
```{r bart, fig.width = 8, fig.height = 4}
bart_fit_std = bart(
  x.train = x.train_std,
  y.train = y.train_std,
  verbose = FALSE, keeptrees = TRUE)

bart_vi_std = as.data.frame(bart_fit_std$varcount) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "avg") 

pred_std = predict(bart_fit_std, newdata = x.test_std)
bart_predictions_std = as.data.frame(t(apply(pred_std, 2, quantile, probs = c(0.025, 0.975))))
colnames(bart_predictions_std) = c("ci_low", "ci_high")

bart_predictions_std$estimate = apply(pred_std, 2, mean) 
bart_predictions_std$observed = y.test_std

coverage_95_bart_std = mean(
  bart_predictions_std$observed >= bart_predictions_std$ci_low &
  bart_predictions_std$observed <= bart_predictions_std$ci_high,
  na.rm = TRUE
)

ggplot(bart_predictions_std, aes(x = estimate, y = observed)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, alpha = 0.35) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted with 95% Prediction Intervals (season)", 
       subtitle = sprintf("95%% coverage = %.1f%% (n=%d)", 100*coverage_95_bart_std, nrow(bart_predictions_std))) +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))
```

### New metrics for standard
```{r error}
pred_mean = bart_predictions_std$estimate
obs = bart_predictions_std$observed
n = length(obs)
p = ncol(x.train_std) 

R2 = 1 - sum((obs - pred_mean)^2) / sum((obs - mean(obs))^2)
adj_R2 = 1 - (1 - R2) * (n - 1) / (n - p - 1)
RMSE = sqrt(mean((obs - pred_mean)^2))
MAE = mean(abs(obs - pred_mean))
coverage_95 = coverage_95_bart_std

bart_metrics = tibble(
  R2 = R2,
  adj_R2 = adj_R2,
  RMSE = RMSE,
  MAE = MAE,
  coverage_95 = coverage_95
)

knitr::kable(bart_metrics, caption = "Error metrics for BART")
```

### Socieconomic BART
```{r bart_fit_soc, fig.width = 8, fig.height = 4}
bart_fit_soc = bart(
  x.train = x.train_soc,
  y.train = y.train_soc,
  verbose = FALSE, keeptrees = TRUE)

bart_vi_soc = as.data.frame(bart_fit_soc$varcount) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "avg") 

pred_soc = predict(bart_fit_soc, newdata = x.test_soc)
bart_predictions_soc = as.data.frame(t(apply(pred_soc, 2, quantile, probs = c(0.025, 0.975))))
colnames(bart_predictions_soc) = c("ci_low", "ci_high")

bart_predictions_soc$estimate = apply(pred_soc, 2, mean) 
bart_predictions_soc$observed = y.test_soc

coverage_95_bart_soc = mean(
  bart_predictions_soc$observed >= bart_predictions_soc$ci_low &
  bart_predictions_soc$observed <= bart_predictions_soc$ci_high,
  na.rm = TRUE
)

ggplot(bart_predictions_soc, aes(x = estimate, y = observed)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, alpha = 0.35) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted with 95% Prediction Intervals (season, socioeconomic)", 
       subtitle = sprintf("95%% coverage = %.1f%% (n=%d)", 100*coverage_95_bart_soc, nrow(bart_predictions_soc))) +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))
```

```{r}
pred_mean_soc = bart_predictions_soc$estimate
obs_soc = bart_predictions_soc$observed
n_soc = length(obs_soc)
p_soc = ncol(x.train_soc)

R2_soc = 1 - sum((obs_soc - pred_mean_soc)^2) / sum((obs_soc - mean(obs_soc))^2)
adj_R2_soc = 1 - (1 - R2_soc) * (n_soc - 1) / (n_soc - p_soc - 1)
RMSE_soc = sqrt(mean((obs_soc - pred_mean_soc)^2))
MAE_soc = mean(abs(obs_soc - pred_mean_soc))

bart_metrics_soc = tibble(
  R2 = R2_soc,
  adj_R2 = adj_R2_soc,
  RMSE = RMSE_soc,
  MAE = MAE_soc,
  coverage_95 = coverage_95_bart_soc
)

knitr::kable(bart_metrics_soc, caption = "Error metrics for BART (socieconomic)")
```

### 99.7% intervals for both BART models
```{r}
probs = c(0.00135, 0.99865)

bart_predictions_std_97 = as.data.frame(t(apply(pred_std, 2, quantile, probs = probs)))
colnames(bart_predictions_std_97) = c("ci_low", "ci_high")

bart_predictions_std_97$estimate = apply(pred_std, 2, mean) 
bart_predictions_std_97$observed = y.test_std

coverage_97_bart_std = mean(
  bart_predictions_std_97$observed >= bart_predictions_std_97$ci_low &
  bart_predictions_std_97$observed <= bart_predictions_std_97$ci_high,
  na.rm = TRUE
)

bart_predictions_soc_97 = as.data.frame(t(apply(pred_soc, 2, quantile, probs = probs)))
colnames(bart_predictions_soc_97) = c("ci_low", "ci_high")

bart_predictions_soc_97$estimate = apply(pred_soc, 2, mean) 
bart_predictions_soc_97$observed = y.test_soc

coverage_97_bart_soc = mean(
  bart_predictions_soc_97$observed >= bart_predictions_soc_97$ci_low &
  bart_predictions_soc_97$observed <= bart_predictions_soc_97$ci_high,
  na.rm = TRUE
)
```

```{r, fig.width = 8, fig.height = 8}
std = ggplot(bart_predictions_std_97, aes(x = estimate, y = observed)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, alpha = 0.35) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted with 99.7% Prediction Intervals (season)", 
       subtitle = sprintf("95%% coverage = %.1f%% (n=%d)", 100*coverage_97_bart_std, nrow(bart_predictions_std))) +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))

soc = ggplot(bart_predictions_soc_97, aes(x = estimate, y = observed)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, alpha = 0.35) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted with 99.7% Prediction Intervals (socioeconomic, season)", 
       subtitle = sprintf("95%% coverage = %.1f%% (n=%d)", 100*coverage_97_bart_soc, nrow(bart_predictions_soc))) +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))

std / soc
```

# Investigating row count mishaps
Looks like I have 833 rows? I have that amount of observations for some reason. Let me check. Alright, socieconomic had duplicate HSA IDs because counties can share HSA IDs. You can run this after, I saved the data to a new file. 

```{r}
nrow(df_complete)
```

```{r}
overall2 = read.csv("../forecasting_metrics_overall.csv")
pop_desc = read.csv("../us_hsa_county_popdesc.csv")
socioecon = read.csv("../Report 7/socioecon.csv") 

us_hsa_county_popdesc_bygroup = pop_desc %>%
  select(state, hsa_nci_id, area_km2_hsa, area_km2_state, n_hsa) %>% 
  distinct()

overall = overall2 %>%
  left_join(us_hsa_county_popdesc_bygroup, by = c("state", "hsa_nci_id"))

df_var = overall %>% 
  filter(season == "2023/24", horizon == 3) %>%
  distinct()
```

```{r}
socioecon_unique = socioecon %>%
  group_by(hsa_nci_id) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

```{r}
df_complete = left_join(df_var, socioecon_unique, by = "hsa_nci_id") %>% mutate(density_ratio = density_hsa / density_state, area_ratio = area_km2_hsa / area_km2_state)
```

```{r}
write.csv(df_complete, "../complete_metrics.csv", row.names = FALSE)
```

```{r}
df_complete = read.csv("../complete_metrics.csv")
df_complete = drop_na(df_complete)
```


