---
title: "Report 3"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=4, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(patchwork)
library(zoo)
```

# Introduction
This week, I will try to find normal/average values for DTW metrics for all HSA IDs. After that, I'll go back and contextualize the metrics (with examples) and find outlier DTW metric values and investigate their trends and if those are captured well. Lastly, I'll explain the metrics more thoroughly (including how they are defined and implemented as algorithsm - for both DTW and Frechet) and implement unwarped amplitude since DTW tends to warp amplitude and then compare which doesn't seem useful.

## Data processing
```{r}
hsa_state = read.csv("../hsa_state_inc.csv")
```

## Metric functions
```{r}
discrete_frechet = function(P, Q, dist_fun = NULL) {
  # P: n x d matrix or data.frame
  # Q: m x d matrix or data.frame
  # dist_fun: function(x, y) returning scalar distance
  
  if (is.null(dist_fun)) {
    dist_fun = function(x, y) sqrt(sum((x - y)^2))
  }
  
  P = as.matrix(P)
  Q = as.matrix(Q)
  
  n = nrow(P)
  m = nrow(Q)
  
  # distance matrix
  dist_mat = matrix(0, n, m)
  for (i in 1:n)
    for (j in 1:m)
      dist_mat[i, j] = dist_fun(P[i, ], Q[j, ])
  
  # DP table
  D = matrix(Inf, n, m)
  Prev = array(NA_integer_, dim = c(n, m, 2))
  
  # initialization
  D[1, 1] = dist_mat[1, 1]
  
  for (i in 2:n) {
    D[i, 1] = max(D[i - 1, 1], dist_mat[i, 1])
    Prev[i, 1, ] = c(i - 1, 1)
  }
  
  for (j in 2:m) {
    D[1, j] = max(D[1, j - 1], dist_mat[1, j])
    Prev[1, j, ] = c(1, j - 1)
  }
  
  # DP recursion
  for (i in 2:n) {
    for (j in 2:m) {
      candidates = c(
        D[i - 1, j],
        D[i, j - 1],
        D[i - 1, j - 1]
      )
      
      k = which.min(candidates)
      
      if (k == 1) Prev[i, j, ] = c(i - 1, j)
      if (k == 2) Prev[i, j, ] = c(i, j - 1)
      if (k == 3) Prev[i, j, ] = c(i - 1, j - 1)
      
      D[i, j] = max(dist_mat[i, j], candidates[k])
    }
  }
  
  # backtrack optimal path
  i = n; j = m
  path = matrix(c(i, j), ncol = 2)
  
  while (!(i == 1 && j == 1)) {
    p = Prev[i, j, ]
    i = p[1]; j = p[2]
    path = rbind(c(i, j), path)
  }
  
  # stepwise Frechet values
  step_dist = apply(path, 1, function(idx)
    dist_mat[idx[1], idx[2]]
  )
  
  frechet_steps = cummax(step_dist)
  
  list(
    frechet = D[n, m],
    path = path,
    step_distance = step_dist,
    frechet_by_step = frechet_steps
  )
}

dtw_distance = function(P, Q, dist_fun = NULL) {
    # P, Q: numeric vectors or matrices (rows = time
    # points)
    if (is.null(dist_fun))
        dist_fun = function(x, y) sqrt(sum((x - y)^2))

    P = as.matrix(P)
    Q = as.matrix(Q)

    n = nrow(P)
    m = nrow(Q)

    # distance matrix
    dist_mat = matrix(0, n, m)
    for (i in 1:n) for (j in 1:m) dist_mat[i, j] = dist_fun(P[i,
        ], Q[j, ])

    # cumulative cost matrix
    D = matrix(Inf, n, m)
    D[1, 1] = dist_mat[1, 1]

    # initialize first row/column
    for (i in 2:n) D[i, 1] = dist_mat[i, 1] + D[i - 1, 1]
    for (j in 2:m) D[1, j] = dist_mat[1, j] + D[1, j - 1]

    # fill rest of matrix
    for (i in 2:n) {
        for (j in 2:m) {
            D[i, j] = dist_mat[i, j] + min(D[i - 1, j], D[i,
                j - 1], D[i - 1, j - 1])
        }
    }

    # backtrack optimal path
    i = n
    j = m
    path = matrix(c(i, j), ncol = 2)
    while (!(i == 1 && j == 1)) {
        if (i == 1) {
            j = j - 1
        } else if (j == 1) {
            i = i - 1
        } else {
            # choose minimum predecessor
            k = which.min(c(D[i - 1, j - 1], D[i - 1, j], D[i,
                j - 1]))
            if (k == 1) {
                i = i - 1
                j = j - 1
            } else if (k == 2) {
                i = i - 1
            } else {
                j = j - 1
            }
        }
        path = rbind(c(i, j), path)
    }

    # stepwise cumulative cost along path
    step_cost = dist_mat[path]
    cum_cost = cumsum(step_cost)

    list(dtw_distance = D[n, m], path = path, step_cost = step_cost,
        cum_cost = cum_cost)
}
```

```{r, fig.height = 10, fig.width = 7}
graph_hsa = function(hsa_id = 688) {
  hsa = hsa_state[hsa_state$hsa_nci_id == hsa_id, ]
  
  P = as.matrix(hsa$inc_hsa)
  Q = as.matrix(hsa$inc_state)
  
  res_dtw = dtw_distance(P, Q)
  
  res = discrete_frechet(P, Q)
  
  normal = ggplot(hsa) + geom_line(linewidth = 1, (aes(x = seq_len(148),
      y = inc_hsa, color = "HSA"))) + geom_line(linewidth = 1,
      (aes(x = seq_len(148), y = inc_state, color = "State"))) +
      labs(x = "index (weeks, 2022 Week 39 to 2025 Week 30)", y = "infected_value (?)",
          title = paste0("Local vs. state inc values for HSA ID ", hsa_id, " from ", hsa$state)) +
      theme_minimal() + scale_color_manual(name = "Legend", values = c(HSA = "blue",
      State = "red"))
  
  frechet = ggplot() + geom_line(aes(x=seq_len(length(res$frechet_by_step)), y=res$frechet_by_step), linewidth = 1) + 
    theme_minimal() + labs(x="Steps", y="Frechet distance", title = "Frechet distance by step")
  
  dtw = ggplot() + geom_line(aes(x = seq_len(length(res_dtw$cum_cost)), y = res_dtw$cum_cost), linewidth = 1) + 
    theme_minimal() + labs(x = "Steps", y = "Cumulative DTW cost", title = "DTW cumulative cost along optimal path by step")
  
  # dtw adjusted
  P_warped = rep(NA, length(Q))
  i = res_dtw$path[,1]
  j = res_dtw$path[,2]
  P_warped[j] = P[i]
  #P_warped = na.approx(P_warped, rule = 2)
  
  df_plot = data.frame(
    index = seq_len(length(Q)),
    HSA_aligned = P_warped,
    State = Q
  )
  
  warped = ggplot(df_plot) +
    geom_line(aes(x = index, y = HSA_aligned, color = "HSA (warped)"), linewidth = 1) +
    geom_line(aes(x = index, y = State, color = "State"), linewidth = 1) +
    labs(
      x = "index (weeks, 2022 Week 39 to 2025 Week 30)",
      y = "infected_value (?)",
      title = paste0("DTW-aligned Local vs. state inc values for HSA ID ", hsa_id, " from ", hsa$state)
    ) +
    theme_minimal() +
    scale_color_manual(name = "Legend", values = c("HSA (warped)" = "blue", State = "red"))
  
  (normal / warped) / (frechet + dtw)
}
```

```{r}
dtw_stats = function(hsa_id = 688) {
  hsa = hsa_state[hsa_state$hsa_nci_id == hsa_id, ]
  
  P = as.matrix(hsa$inc_hsa)
  Q = as.matrix(hsa$inc_state)
  
  res_dtw = dtw_distance(P, Q)
  
  res = discrete_frechet(P, Q)
  
  i = res_dtw$path[,1]
  j = res_dtw$path[,2]

  delta_t = j - i
  delta_x = P[i,1] - Q[j,1]

  stats_list = list(
    mean_abs_lag = mean(abs(delta_t)),
    signed_lag   = mean(delta_t),
    lag_sd       = sd(delta_t),
    dtw_mae      = mean(abs(delta_x)),
    dtw_rmse     = sqrt(mean(delta_x^2)),
    rmse = sqrt(mean((P[,1] - Q[,1])^2)),
    amp_ratio    = median(P[,1] / Q[,1], na.rm=TRUE),
    dtw_amp_ratio    = median(P[i,1] / Q[j,1], na.rm=TRUE),
    lag_amp_corr = cor(abs(delta_t), abs(delta_x))
  )

  stats = data.frame(
    Metric = names(stats_list),
    Value  = unname(unlist(stats_list))
  )
  
  return(stats)
}
```

```{r, fig.height = 14, fig.width = 10}
# set.seed(77)
id_list = hsa_state$hsa_nci_id
id = sample(id_list, 1)
graph_hsa(hsa_id = id)
knitr::kable(dtw_stats(hsa_id = id), digits = 3)  
```

# All-HSA metrics
I want to initially make two tables, one that holds HSA IDs and another table that holds metrics and statistics about their distributions from all the HSA IDs.

I tried running this with all 98,295 HSA IDs and that went terribly (took literal eons) so I'm going to use stratified sampling by state and go and get 10,000 IDs.
```{r}
set.seed(77)

hsa_unique = hsa_state %>%
  distinct(hsa_nci_id, state)

state_sizes = hsa_unique %>%
  count(state, name = "N_state") %>%
  mutate(n_sample = ceiling(100 * N_state / length(hsa_unique$hsa_nci_id)))

hsa_list = list()
for (state_name in unique(hsa_state$state)) {
  state_subset = hsa_unique %>% filter(state == state_name)
  state_temp = sample(state_subset$hsa_nci_id, state_sizes[state_sizes$state == state_name, "n_sample"])
  hsa_list = append(hsa_list, state_temp)
}
```

```{r, cache = TRUE}
results_list = vector("list", length(hsa_list$hsa_nci_id))

for (i in seq_along(hsa_list$hsa_nci_id)) {
  ID = hsa_list[[i]]
  
  out = pivot_wider(as.data.frame(dtw_stats(hsa_id = ID)), names_from = Metric, values_from = Value)
  out$ID = ID
  out = out[, c("ID", setdiff(names(out), "ID"))]

}

dtw_results <- do.call(rbind, results_list)
```

```{r}
res = vector("list", length(hsa_list))

for (i in seq_along(hsa_list)) {
  ID = hsa_list[[i]]
  
  out = pivot_wider(as.data.frame(dtw_stats(hsa_id = ID)), names_from  = Metric, values_from = Value)
  
  out$ID = ID
  out = out[, c("ID", setdiff(names(out), "ID"))]
  
  res[[i]] = out
}

dtw_results = do.call(rbind, res)
```

```{r}
dtw_results
knitr::kable(head(dtw_results))
```









































