---
title: "Report 6"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=4, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(stringr)
library(lubridate)
library(patchwork)
library(mosaic)
library(tidyr)
library(formatR)

library(data.table)
library(tidycensus)
library(dbarts)
library(iml)
```

# Introduction
This week, we are going to introduce new socioeconomic variables 

# Data processing
```{r}
v23 = load_variables(2023, "acs5", cache = TRUE)
```

## Education
```{r}
education_county = get_acs(geography = "county", variables = c(total = "B15003_001", bachelors = "B15003_021", high_school = "B15003_017"), year = 2023) %>%
  select(GEOID, NAME, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>% mutate (state_ID = as.numeric(GEOID) %/% 1000, high_school_pct = high_school / total)

head(education_county)
```

I think we choose only high school here because it has way more correlation and people who go to college usually leave their county.

```{r}
education_county = education_county[,c(1:2, 6:7)]
```

Now, we get the state-level version.
```{r}
education_state = get_acs(geography = "state", variables = c(total = "B15003_001", high_school = "B15003_017"), year = 2023) %>% select(GEOID, NAME, variable, estimate) %>% pivot_wider(names_from = variable, values_from = estimate) %>% mutate (high_school_pct_state = high_school / total)
```

```{r}
education_state$state_ID = as.numeric(education_state$GEOID)
```

```{r}
education = left_join(education_county, education_state[,c(5:6)], by = "state_ID") %>% mutate(education_diff = high_school_pct_state - high_school_pct) %>% select(GEOID, state_ID, NAME, education_diff)

head(education)
```
The difference will be state minus county going forward as well.

## Income
```{r}
income_county = get_acs(geography = "county", variable = "B19013_001", year = 2023) %>% select(GEOID, NAME, variable, estimate) %>% mutate(state_ID = as.numeric(GEOID) %/% 1000)

income_state = get_acs(geography = "state", variable = "B19013_001", year = 2023) %>% select(GEOID, NAME, variable, estimate) %>% mutate(state_ID = as.numeric(education_state$GEOID), state_estimate = estimate)
```

```{r}
income = left_join(income_county, income_state[,c(5:6)], by = "state_ID") %>% mutate(income_diff = log(state_estimate) - log(estimate)) %>% select(GEOID, state_ID, NAME, income_diff)

head(income)
```

## Poverty level
```{r}
poverty_county = get_acs(geography = "county", variables = c(total = "B17001_001", poverty = "B17001_002"), year = 2023) %>% select(-moe) %>% pivot_wider(names_from = variable, values_from = estimate) %>% mutate (state_ID = as.numeric(GEOID) %/% 1000, poverty_pct = poverty / total)

poverty_state = get_acs(geography = "state", variables = c(total = "B17001_001", poverty = "B17001_002"), year = 2023) %>% select(-moe) %>% pivot_wider(names_from = variable, values_from = estimate) %>% mutate (state_ID = as.numeric(GEOID), poverty_pct_state = poverty / total)
```

```{r}
poverty = left_join(poverty_county, poverty_state[,c(5:6)], by = "state_ID") %>% mutate(poverty_diff = poverty_pct_state - poverty_pct) %>% select(GEOID, state_ID, NAME, poverty_diff)

head(poverty)
```

## Unemployment rate
```{r}
unemployment_county = get_acs(geography = "county", variables = c(total = "B23025_003", unemployed = "B23025_005"), year = 2023) %>% select(-moe) %>% pivot_wider(names_from = variable, values_from = estimate) %>% mutate (state_ID = as.numeric(GEOID) %/% 1000, unemployment_pct = unemployed / total)

unemployment_state = get_acs(geography = "state", variables = c(total = "B23025_003", unemployed = "B23025_005"), year = 2023) %>% select(-moe) %>% pivot_wider(names_from = variable, values_from = estimate) %>% mutate (state_ID = as.numeric(GEOID), unemployment_pct_state = unemployed / total)
```

```{r}
unemployment = left_join(unemployment_county, unemployment_state[,c(5:6)], by = "state_ID") %>% mutate(unemployment_diff = unemployment_pct_state - unemployment_pct)  %>% select(GEOID, state_ID, NAME, unemployment_diff)

head(unemployment)
```

## Socioeconomic data
```{r}
hsa_list = readxl::read_xls("Health.Service.Areas.xls")
head(hsa_list)
```

```{r}
socioecon = left_join(education, income[,c(1,4)], by = "GEOID") %>% left_join(poverty[,c(1,4)], by = "GEOID") %>% left_join(unemployment[,c(1,4)], by = "GEOID") %>% select(-state_ID)

colnames(hsa_list) = c("hsa_nci_id", "desc", "state-county", "GEOID")
socioecon = left_join(socioecon, hsa_list[,c(1,4)], by = "GEOID") %>% select(GEOID, hsa_nci_id, education_diff, income_diff, poverty_diff, unemployment_diff, NAME)

head(socioecon)
```
```{r}
write.csv(socioecon, file = "socioecon.csv", row.names = FALSE)
```

# Training BART
## Data preprocessing
```{r}
overall2 = read.csv("../forecasting_metrics_overall.csv")
pop_desc = read.csv("../us_hsa_county_popdesc.csv")
socioecon = read.csv("socioecon.csv") 

us_hsa_county_popdesc_bygroup = pop_desc %>%
  select(state, hsa_nci_id, area_km2_hsa, area_km2_state, n_hsa) %>% 
  distinct()

overall = overall2 %>%
  left_join(us_hsa_county_popdesc_bygroup, by = c("state", "hsa_nci_id"))

df_var = overall %>% 
  filter(season == "2023/24", horizon == 3) %>%
  distinct()
```

```{r}
df_complete = left_join(df_var, socioecon, by = "hsa_nci_id") %>% mutate(density_ratio = density_hsa / density_state, area_ratio = area_km2_hsa / area_km2_state)
```

```{r}
write.csv(df_complete, "../complete_metrics.csv", row.names = FALSE)
```

Okay, there are two HSAs with missing income diff values and I am dropping them. They just don't have HSA level income values all the way back from the income census data. 

```{r}
df_complete = read.csv("../complete_metrics.csv")
df_complete = drop_na(df_complete)
```

## Training
### Train and test split
```{r}
set.seed(123)
n = nrow(df_complete)
train_idx = sample(seq_len(n), size = floor(0.8*n)) # 80% train
test_idx = setdiff(seq_len(n), train_idx)

vars_std = c("pop_ratio", "pct_urban", "density_ratio", "n_hsa", "area_ratio")

x.train_std = df_complete[train_idx, vars_std]
y.train_std = df_complete$diff_wis_season[train_idx]

x.test_std = df_complete[test_idx, vars_std]
y.test_std = df_complete$diff_wis_season[test_idx]

vars_soc = c("pop_ratio", "pct_urban", "density_ratio", "n_hsa", "area_ratio", "education_diff", "income_diff", "poverty_diff", "unemployment_diff")

x.train_soc = df_complete[train_idx, vars_soc]
y.train_soc = df_complete$diff_wis_season[train_idx]

x.test_soc = df_complete[test_idx, vars_soc]
y.test_soc = df_complete$diff_wis_season[test_idx]
```

```{r bart_fit_std}
bart_fit_std = bart(
  x.train = x.train_std,
  y.train = y.train_std,
  verbose = FALSE, keeptrees = TRUE)

bart_vi_std = as.data.frame(bart_fit_std$varcount) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "avg") 

pred_std = predict(bart_fit_std, newdata = x.test_std)
bart_predictions_std = as.data.frame(t(apply(pred_std, 2, quantile, probs = c(0.025, 0.975))))
colnames(bart_predictions_std) = c("ci_low", "ci_high")

bart_predictions_std$estimate = apply(pred_std, 2, mean) 
bart_predictions_std$observed = y.test_std

coverage_95_bart_std = mean(
  bart_predictions_std$observed >= bart_predictions_std$ci_low &
  bart_predictions_std$observed <= bart_predictions_std$ci_high,
  na.rm = TRUE
)
```

```{r bart_fit_soc}
bart_fit_soc = bart(
  x.train = x.train_soc,
  y.train = y.train_soc,
  verbose = FALSE, keeptrees = TRUE)

bart_vi_soc = as.data.frame(bart_fit_soc$varcount) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "avg") 

pred_soc = predict(bart_fit_soc, newdata = x.test_soc)
bart_predictions_soc = as.data.frame(t(apply(pred_soc, 2, quantile, probs = c(0.025, 0.975))))
colnames(bart_predictions_soc) = c("ci_low", "ci_high")

bart_predictions_soc$estimate = apply(pred_soc, 2, mean) 
bart_predictions_soc$observed = y.test_soc

coverage_95_bart_soc = mean(
  bart_predictions_soc$observed >= bart_predictions_soc$ci_low &
  bart_predictions_soc$observed <= bart_predictions_soc$ci_high,
  na.rm = TRUE
)
```

```{r feature_imp, fig.width = 8, fig.height=6}
std_feature = ggplot(bart_vi_std) + geom_col(aes(y = reorder(variable, avg), x = avg), fill='skyblue', col='black') + labs(y = "Feature", x = "Importance", title = "Feature importance (in-built, season)") + coord_cartesian(xlim = c(0, 90))

soc_feature = ggplot(bart_vi_soc) + geom_col(aes(y = reorder(variable, avg), x = avg), fill='skyblue', col='black') + labs(y = "Feature", x = "Importance", title = "Feature importance (in-built, socioeconomic, season)") + coord_cartesian(xlim = c(0,90))

std_feature / soc_feature
```

```{r, fig.width = 8, fig.height = 8}
std = ggplot(bart_predictions_std, aes(x = estimate, y = observed)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, alpha = 0.35) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted with 95% Prediction Intervals (season)", 
       subtitle = sprintf("95%% coverage = %.1f%% (n=%d)", 100*coverage_95_bart_std, nrow(bart_predictions_std))) +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))

soc = ggplot(bart_predictions_soc, aes(x = estimate, y = observed)) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, alpha = 0.35) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted diff_wis_season", y = "Observed diff_wis_season",
       title = "Observed vs Predicted with 95% Prediction Intervals (socioeconomic, season)", 
       subtitle = sprintf("95%% coverage = %.1f%% (n=%d)", 100*coverage_95_bart_soc, nrow(bart_predictions_soc))) +
  theme_minimal() + scale_x_continuous(breaks=seq(-1,2,by=0.25)) + coord_cartesian(xlim = c(-0.25, 1))

std / soc
```
# SHAP feature importance 
## Standard
```{r shap_std, cache=TRUE}
predictor_std = Predictor$new(
  model = bart_fit_std,
  data  = as.data.frame(df_complete[, vars_std]),
  y     = df_complete$diff_wis_season,
  predict.function = function(object, newdata) {
    colMeans(predict(object, newdata = newdata))
  }
)

all_shap_std = rbindlist(lapply(1:nrow(df_complete), function(i) {
  sh = Shapley$new(
    predictor_std,
    x.interest = df_complete[i, vars_std],
    sample.size   = 50 
  )
  out = sh$results
  out$obs = i        
  out
}))
```

```{r std_shap_table}
global_imp_std = all_shap_std[, .(mean_abs_phi = mean(abs(phi))), by = feature]
knitr::kable(global_imp_std[order(-mean_abs_phi)], caption = "(Season) Features by importance (descending)")
```

This graph basically discusses how much "movement" ($|\phi|$) each feature contributes in turning the baseline into the estimate. 
```{r std_shap_graph, fig.width = 10}
ggplot(global_imp_std, aes(x = reorder(feature, mean_abs_phi), y = mean_abs_phi)) +
  geom_col(fill='skyblue', col='black') +
  coord_flip() +
  labs(x = "Feature", y = "Mean of absolute value of SHAP",
       title = "(Season) Global SHAP feature importance")

```

## Socieconomic
```{r shap_soc, cache=TRUE}
predictor_soc = Predictor$new(
  model = bart_fit_soc,
  data  = as.data.frame(df_complete[, vars_soc]),
  y     = df_complete$diff_wis_season,
  predict.function = function(object, newdata) {
    colMeans(predict(object, newdata = newdata))
  }
)

all_shap_soc = rbindlist(lapply(1:nrow(df_complete), function(i) {
  sh = Shapley$new(
    predictor_soc,
    x.interest = df_complete[i, vars_soc],
    sample.size   = 50 
  )
  out = sh$results
  out$obs = i        
  out
}))
```

```{r soc_shap_table}
global_imp_soc = all_shap_soc[, .(mean_abs_phi = mean(abs(phi))), by = feature]
knitr::kable(global_imp_soc[order(-mean_abs_phi)], caption = "(Season, Socioeconomic) Features by importance (descending)")
```

This graph basically discusses how much "movement" ($|\phi|$) each feature contributes in turning the baseline into the estimate. 
```{r soc_shap_graph, fig.width = 10}
ggplot(global_imp_soc, aes(x = reorder(feature, mean_abs_phi), y = mean_abs_phi)) +
  geom_col(fill='skyblue', col='black') +
  coord_flip() +
  labs(x = "Feature", y = "Mean of absolute value of SHAP",
       title = "(Season, Socioeconomic) Global SHAP feature importance")

```





