---
title: "Report 5"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=4, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(patchwork)
library(zoo)
library(kableExtra)
library(corrplot)
library(forcats)
```

# Introduction
This week, I will analyze the two FluSight models I have shortlisted and take some notes (which will be stored here). I will try to find some real trends (maybe with linear regression or a deeper analysis of that correlation plot) to look at DTW metrics in more detail. I also want to look at how trends happen over HSAs (? possibly because then I could take a look at some outliers) and look at how trends in states actually do different things. Also also, I think I can look at every row (HSA ID) now because the code is more optimized so that's nice. Looking forward to it!

# Data processing
```{r}
hsa_state = read.csv("../hsa_state_inc.csv")
```

## Metric functions
```{r}
source("../metric_functions.R")
source("../graphstat_functions.R")
source("../write_dtw.R")
```

```{r, fig.height = 14, fig.width = 10}
# set.seed(77)
# id_list = hsa_state$hsa_nci_id
# id = sample(id_list, 1)
# graph_hsa(hsa_id = id)
# knitr::kable(dtw_stats(hsa_id = id), digits = 3)  
```

```{r, cache = TRUE}
dtw_results = read_csv("../all_dtw_results.csv")
hsa_dtw = dtw_results |> left_join(hsa_state, by = join_by(ID == hsa_nci_id)) 

metrics = setdiff(names(dtw_results), "ID")
summary = reshape(do.call(rbind, lapply(metrics, \(m) data.frame(Metric = m, Var2 = names(summary(dtw_results[[m]])), Freq = as.numeric(summary(dtw_results[[m]])), row.names = NULL))), idvar = "Metric", timevar = "Var2", direction = "wide")
colnames(summary) = c("Metric", "Minimum", "1st Quartile", "Median", "Mean", "3rd Quartile", "Maximum")
rownames(summary) = NULL

knitr::kable(summary |> mutate(across(where(is.numeric), ~ round(., 3))), caption = "Summary Statistics for Dynamic Time Warping Metrics (+ Frechet) on HSA IDs sampled with stratification (state)") 
```

# FluSight models
## Cornell-JHU 
### Notes
The model is a hybrid SIR - Bayesian hierarchical discrepancy model for infectious disease forecasting. After a quick run-through, I couldn't find much more about the specifics of their model but this is the type of model. 

### Implementation
I think this one will be easier to implement, might have to transition to Python or another language that they use

## CMU
### Notes
The model is a AR model with specific data that CMU externally sources, I think. I found another link with more (https://zoltardata.com/model/307)

### Implementation
I'll try to first implement a basic AR model and then see if I can get access to their data.

# More plots (with interpretations)
## Analysis of last week's corerlation plot
## Trends per state 
```{r, fig.length = 12, fig.width = 12}
dtw_results_long = dtw_results |> pivot_longer(cols = -ID, names_to = "metric", values_to = "value")

id_state_lookup = hsa_dtw |> select(ID, state) |> distinct() 

id_state_lookup$ID = as.numeric(id_state_lookup$ID)

dtw_results_long = dtw_results_long |>
  left_join(id_state_lookup, by = "ID")

# state_counts = id_state_lookup %>%
#   group_by(state) %>%
#   summarise(n = n())
```

```{r, fig.width = 8, fig.height = 5}
ggplot(dtw_results_long, aes(x = value)) + geom_density(fill = "skyblue", alpha = 0.8) + facet_wrap(~ metric, scales = "free", nrow = 2) + 
  theme_bw() + labs(title = "Density Plots for All DTW Metrics (+ Frechet)", x = "Value", y = "Density")
```

Remember from the last report that lag_amp_corr is the correlation between absolute time lag and amplitude warped difference where positive values suggest larger timing shifts coincide with larger amplitude differences while negative values suggest that larger timing shifts coincide with smaller amplitude differences. ANALYSIS COMING SOON.
```{r, fig.height = 15, fig.width = 12}
dtw_lagamp_corr =
  dtw_results_long |>
  filter(metric == "lag_amp_corr")

state_order =
  dtw_lagamp_corr |>
  group_by(state) |>
  summarise(med = median(value, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(med)) |>
  pull(state)

dtw_lagamp_corr$state = factor(dtw_lagamp_corr$state, levels = state_order)

state_counts =
  dtw_lagamp_corr |>
  count(state)

ggplot(dtw_lagamp_corr, aes(x = "", y = value)) + geom_violin(fill = "royalblue",
    alpha = 0.8) + geom_boxplot(fill = "firebrick", alpha = 0.5,
    width = 0.3) + facet_wrap(~state, nrow = 5) + theme_bw() +
    labs(title = "Violin Plots for Lag/Amplitude Correlation",
        x = NULL, y = "Value") + geom_text(data = state_counts,
    aes(x = 1, y = -Inf, label = paste0("n=", n)), vjust = -0.5,
    inherit.aes = FALSE, size = 3)
```

In the plot below, the positive values indicate that DTW improved the RMSE of that HSA over the designated timeframe.
```{r, fig.height = 15, fig.width = 12}
dtw_results_diff =
  hsa_dtw |> mutate(dtw_diff = (rmse - dtw_rmse))

state_order = 
  dtw_results_diff |>
  group_by(state) |>
  summarise(med = median(dtw_diff, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(med)) |>
  pull(state)

dtw_results_diff$state = factor(dtw_results_diff$state, levels = state_order)

ggplot(dtw_results_diff, aes(x = "", y = dtw_diff)) +
 geom_violin(fill = "royalblue", alpha = 0.8) + geom_boxplot(fill = "firebrick", alpha = 0.5, width = 0.3) + theme_bw() + facet_wrap(~state, nrow = 5) +
 labs(title = "Violin Plots for RMSE minus DTW RMSE", x = "Value", y = "Density") + geom_text(data = state_counts, aes(x = 1, y = min(dtw_results_diff$rmse - dtw_results_diff$dtw_rmse) - 0.05, label = paste0("n=", n)), inherit.aes = FALSE, size = 3) 
```

I feel like the violin plots are very hard to interpret for this variable difference.
```{r, fig.height = 12, fig.width = 14}
dtw_results_diff_amp =
  hsa_dtw |> mutate(dtw_diff_amp = (amp_ratio - dtw_amp_ratio))

state_order = 
  dtw_results_diff_amp |>
  group_by(state) |>
  summarise(med = median(dtw_diff_amp, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(med)) |>
  pull(state)

dtw_results_diff_amp$state = factor(dtw_results_diff_amp$state, levels = state_order)

ggplot(dtw_results_diff_amp, aes(x = "", y = dtw_diff_amp)) +
 geom_violin(fill = "royalblue", alpha = 0.8) + geom_boxplot(fill = "firebrick", alpha = 0.5, width = 0.3) + theme_bw() + facet_wrap(~state, nrow = 5) +
 labs(title = "Violin Plots for amplitude ratio minus DTW amplitude ratio", x = "Value", y = "Density") + geom_text(data = state_counts, aes(x = 1, y = min(dtw_results_diff_amp$amp_ratio - dtw_results_diff_amp$dtw_amp_ratio) - 0.05, label = paste0("n=", n)), inherit.aes = FALSE, size = 3)
```

Most states have their values on one side of the line, usually the left line, because DTW often improves how close the amplitude ratio is to 1. For example, states like Illinois and Oregon see a significant collection on one side of the line. However, a couple interesting states (with reasonable $n$ values) that have a more "even" distribution are Virginia, Texas, Georgia, Tennessee, Louisiana and West Virginia. One possible correlation with those states is that they are all southern states and states like Minnesota, Iowa, Idaho, Indiana, New York are all northern. However, Kentucky emerges as an exception so maybe not completely north vs. south. For a state to have a more equal distribution on each side of the line, that means that amplitudes were both above and below the state line. More specifically, if points are to the left and up from the red line, that means that the original amplitude ratio was smaller than the DTW (it fit the original from HSA being smaller to being closer). On the other hand, if points are to the right and down from the red line, that means that the original amplitude ratio was larger than the DTW. 

```{r, fig.height = 12, fig.width = 14}
ggplot(hsa_dtw, aes(x = amp_ratio, y = dtw_amp_ratio)) +
  geom_point() + geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red", linewidth = 1) +
  labs(x = "amp_ratio", y = "dtw_amp_ratio", title = "Amplitude Ratios v. DTW Amplitude Ratios") + theme_bw() + facet_wrap(~state, nrow = 5) +
  geom_text(data = state_counts, aes(x = 2.5, y = 0.1, label = paste0("n=", n)), inherit.aes = FALSE, size = 3) +
  scale_y_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1, 1.25), limits = c(0, 1.25)) + scale_x_continuous(breaks = c(0, 0.5, 1, 1.5, 2, 2.5, 3))
```

# Conclusion
Not much state trends abound, but maybe north vs. south is a possible plotline we can hang onto for the future. Visualization is going well for the most part, I think everything is well contextualized. 






























