---
title: "Report 5"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=4, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(patchwork)
library(zoo)
library(kableExtra)
library(corrplot)
library(forcats)
```

# Introduction
This week, I will analyze the two FluSight models I have shortlisted and take some notes (which will be stored here). I will try to find some real trends (maybe with linear regression or a deeper analysis of that correlation plot) to look at DTW metrics in more detail. I also want to look at how trends happen over HSAs (? possibly because then I could take a look at some outliers) and look at how trends in states actually do different things. Also also, I think I can look at every row (HSA ID) now because the code is more optimized so that's nice. Looking forward to it!

# Data processing
```{r}
hsa_state = read.csv("../hsa_state_inc.csv")
```

## Metric functions
```{r}
source("../metric_functions.R")
source("../graphstat_functions.R")
source("../write_dtw.R")
```

```{r, fig.height = 14, fig.width = 10}
# set.seed(77)
# id_list = hsa_state$hsa_nci_id
# id = sample(id_list, 1)
# graph_hsa(hsa_id = id)
# knitr::kable(dtw_stats(hsa_id = id), digits = 3)  
```

```{r, cache = TRUE}
dtw_results = read_csv("../all_dtw_results.csv")
hsa_dtw = dtw_results |> left_join(hsa_state, by = join_by(ID == hsa_nci_id)) 

metrics = setdiff(names(dtw_results), "ID")
summary = reshape(do.call(rbind, lapply(metrics, \(m) data.frame(Metric = m, Var2 = names(summary(dtw_results[[m]])), Freq = as.numeric(summary(dtw_results[[m]])), row.names = NULL))), idvar = "Metric", timevar = "Var2", direction = "wide")
colnames(summary) = c("Metric", "Minimum", "1st Quartile", "Median", "Mean", "3rd Quartile", "Maximum")
rownames(summary) = NULL

knitr::kable(summary |> mutate(across(where(is.numeric), ~ round(., 3))), caption = "Summary Statistics for Dynamic Time Warping Metrics (+ Frechet) on HSA IDs sampled with stratification (state)") 
```

# FluSight models
## Cornell-JHU 
### Notes
The model is a hybrid SIR - Bayesian hierarchical discrepancy model for infectious disease forecasting. After a quick run-through, I couldn't find much more about the specifics of their model but this is the type of model. 

### Implementation
I think this one will be easier to implement, might have to transition to Python or another language that they use

## CMU
### Notes
The model is a AR model with specific data that CMU externally sources, I think. I found another link with more (https://zoltardata.com/model/307)

### Implementation
I'll try to first implement a basic AR model and then see if I can get access to their data.

# More plots (with interpretations)
## Analysis of last week's corerlation plot
## Trends per state 
```{r, fig.length = 12, fig.width = 12}
dtw_results_long = dtw_results |> pivot_longer(cols = -ID, names_to = "metric", values_to = "value")

id_state_lookup = hsa_dtw |> select(ID, state) |> distinct() 

id_state_lookup$ID = as.numeric(id_state_lookup$ID)

dtw_results_long = dtw_results_long |>
  left_join(id_state_lookup, by = "ID")
```

```{r, fig.width = 8, fig.height = 5}
ggplot(dtw_results_long, aes(x = value)) + geom_density(fill = "skyblue", alpha = 0.8) + facet_wrap(~ metric, scales = "free", nrow = 2) + 
  theme_bw() + labs(title = "Density Plots for All DTW Metrics (+ Frechet)", x = "Value", y = "Density")
```

```{r, fig.height = 15, fig.width = 12}
dtw_lagamp_corr = dtw_results_long |> filter(metric == "lag_amp_corr") |> mutate(state = fct_reorder(state, value, .fun = median, .desc = TRUE))

ggplot(dtw_lagamp_corr, aes(x = metric, y = value)) +
  geom_violin(fill = "royalblue", alpha = 0.8) + geom_boxplot(fill = "firebrick", alpha = 0.5, width = 0.3) + theme_bw() + facet_wrap(~state, nrow = 5) +
 labs(title = "Violin Plots for Lag/Amplitude Correlation", x = "Value", y = "Density") 
```

In the plot below, the positive values indicate that DTW improved the RMSE of that HSA over the designated timeframe.
```{r, fig.height = 15, fig.width = 12}
dtw_results_diff = hsa_dtw |> mutate(dtw_diff = (rmse - dtw_rmse)) |> mutate(state = fct_reorder(state, dtw_diff, .fun = median, .desc = TRUE)) 

ggplot(dtw_results_diff, aes(x = "", y = dtw_diff)) +
 geom_violin(fill = "royalblue", alpha = 0.8) + geom_boxplot(fill = "firebrick", alpha = 0.5, width = 0.3) + theme_bw() + facet_wrap(~state, nrow = 5) +
 labs(title = "Violin Plots for RMSE minus DTW RMSE", x = "Value", y = "Density") 
```

```{r, fig.height = 12, fig.width = 14}
ggplot(hsa_dtw, aes(x = amp_ratio, y = dtw_amp_ratio)) +
  geom_point() + geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red", linewidth = 1) + 
  labs(x = "amp_ratio", y = "dtw_amp_ratio", title = "Amplitude Ratios v. DTW Amplitude Ratios") + theme_bw() + facet_wrap(~state, nrow = 5)
#  scale_y_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1, 1.25), limits = c(0, 1.25)) + scale_x_continuous(breaks = c(0, 0.5, 1, 1.5, 2, 2.5, 3)) 
```

## Outlier examination

# Conclusion

































