---
title: "Report 5"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=4, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(patchwork)
library(zoo)
library(kableExtra)
library(corrplot)
```

# Introduction
This week, I will analyze the two FluSight models I have shortlisted and take some notes (which will be stored here). I will try to find some real trends (maybe with linear regression or a deeper analysis of that correlation plot) to look at DTW metrics in more detail. I also want to look at how trends happen over HSAs (? possibly because then I could take a look at some outliers) and look at how trends in states actually do different things. Also also, I think I can look at every row (HSA ID) now because the code is more optimized so that's nice. Looking forward to it!

# Data processing
```{r}
hsa_state = read.csv("../hsa_state_inc.csv")
```

## Metric functions

```{r}
discrete_frechet = function(P, Q, dist_fun = NULL) {
  # P: n x d matrix or data.frame
  # Q: m x d matrix or data.frame
  # dist_fun: function(x, y) returning scalar distance
  
  if (is.null(dist_fun)) {
    dist_fun = function(x, y) sqrt(sum((x - y)^2))
  }
  
  P = as.matrix(P)
  Q = as.matrix(Q)
  
  n = nrow(P)
  m = nrow(Q)
  
  # distance matrix
  dist_mat = matrix(0, n, m)
  for (i in 1:n)
    for (j in 1:m)
      dist_mat[i, j] = dist_fun(P[i, ], Q[j, ])
  
  # DP table
  D = matrix(Inf, n, m)
  Prev = array(NA_integer_, dim = c(n, m, 2))
  
  # initialization
  D[1, 1] = dist_mat[1, 1]
  
  for (i in 2:n) {
    D[i, 1] = max(D[i - 1, 1], dist_mat[i, 1])
    Prev[i, 1, ] = c(i - 1, 1)
  }
  
  for (j in 2:m) {
    D[1, j] = max(D[1, j - 1], dist_mat[1, j])
    Prev[1, j, ] = c(1, j - 1)
  }
  
  # DP recursion
  for (i in 2:n) {
    for (j in 2:m) {
      candidates = c(
        D[i - 1, j],
        D[i, j - 1],
        D[i - 1, j - 1]
      )
      
      k = which.min(candidates)
      
      if (k == 1) Prev[i, j, ] = c(i - 1, j)
      if (k == 2) Prev[i, j, ] = c(i, j - 1)
      if (k == 3) Prev[i, j, ] = c(i - 1, j - 1)
      
      D[i, j] = max(dist_mat[i, j], candidates[k])
    }
  }
  
  # backtrack optimal path
  i = n; j = m
  path = matrix(c(i, j), ncol = 2)
  
  while (!(i == 1 && j == 1)) {
    p = Prev[i, j, ]
    i = p[1]; j = p[2]
    path = rbind(c(i, j), path)
  }
  
  # stepwise Frechet values
  step_dist = apply(path, 1, function(idx)
    dist_mat[idx[1], idx[2]]
  )
  
  frechet_steps = cummax(step_dist)
  
  list(
    frechet = D[n, m],
    path = path,
    step_distance = step_dist,
    frechet_by_step = frechet_steps
  )
}

dtw_distance = function(P, Q, dist_fun = NULL) {
    # P, Q: numeric vectors or matrices (rows = time
    # points)
    if (is.null(dist_fun))
        dist_fun = function(x, y) sqrt(sum((x - y)^2))

    P = as.matrix(P)
    Q = as.matrix(Q)

    n = nrow(P)
    m = nrow(Q)

    # distance matrix
    dist_mat = matrix(0, n, m)
    for (i in 1:n) for (j in 1:m) dist_mat[i, j] = dist_fun(P[i,
        ], Q[j, ])

    # cumulative cost matrix
    D = matrix(Inf, n, m)
    D[1, 1] = dist_mat[1, 1]

    # initialize first row/column
    for (i in 2:n) D[i, 1] = dist_mat[i, 1] + D[i - 1, 1]
    for (j in 2:m) D[1, j] = dist_mat[1, j] + D[1, j - 1]

    # fill rest of matrix
    for (i in 2:n) {
        for (j in 2:m) {
            D[i, j] = dist_mat[i, j] + min(D[i - 1, j], D[i,
                j - 1], D[i - 1, j - 1])
        }
    }

    # backtrack optimal path
    i = n
    j = m
    path = matrix(c(i, j), ncol = 2)
    while (!(i == 1 && j == 1)) {
        if (i == 1) {
            j = j - 1
        } else if (j == 1) {
            i = i - 1
        } else {
            # choose minimum predecessor
            k = which.min(c(D[i - 1, j - 1], D[i - 1, j], D[i,
                j - 1]))
            if (k == 1) {
                i = i - 1
                j = j - 1
            } else if (k == 2) {
                i = i - 1
            } else {
                j = j - 1
            }
        }
        path = rbind(c(i, j), path)
    }

    # stepwise cumulative cost along path
    step_cost = dist_mat[path]
    cum_cost = cumsum(step_cost)

    list(dtw_distance = D[n, m], path = path, step_cost = step_cost,
        cum_cost = cum_cost)
}
```

## Graphing and stats functions
```{r, fig.height = 10, fig.width = 7}
graph_hsa = function(hsa_id = 688) {
  hsa = hsa_state[hsa_state$hsa_nci_id == hsa_id, ]
  
  P = as.matrix(hsa$inc_hsa)
  Q = as.matrix(hsa$inc_state)
  
  res_dtw = dtw_distance(P, Q)
  
  res = discrete_frechet(P, Q)
  
  normal = ggplot(hsa) + geom_line(linewidth = 1, (aes(x = seq_len(148),
      y = inc_hsa, color = "HSA"))) + geom_line(linewidth = 1,
      (aes(x = seq_len(148), y = inc_state, color = "State"))) +
      labs(x = "index (weeks, 2022 Week 39 to 2025 Week 30)", y = "infected_value (?)",
          title = paste0("Local vs. state inc values for HSA ID ", hsa_id, " from ", hsa$state)) +
      theme_minimal() + scale_color_manual(name = "Legend", values = c(HSA = "blue",
      State = "red"))
  
  frechet = ggplot() + geom_line(aes(x=seq_len(length(res$frechet_by_step)), y=res$frechet_by_step), linewidth = 1) + 
    theme_minimal() + labs(x="Steps", y="Frechet distance", title = "Frechet distance by step")
  
  dtw = ggplot() + geom_line(aes(x = seq_len(length(res_dtw$cum_cost)), y = res_dtw$cum_cost), linewidth = 1) + 
    theme_minimal() + labs(x = "Steps", y = "Cumulative DTW cost", title = "DTW cumulative cost along optimal path by step")
  
  # dtw adjusted
  P_warped = rep(NA, length(Q))
  i = res_dtw$path[,1]
  j = res_dtw$path[,2]
  P_warped[j] = P[i]
  #P_warped = na.approx(P_warped, rule = 2)
  
  df_plot = data.frame(
    index = seq_len(length(Q)),
    HSA_aligned = P_warped,
    State = Q
  )
  
  warped = ggplot(df_plot) +
    geom_line(aes(x = index, y = HSA_aligned, color = "HSA (warped)"), linewidth = 1) +
    geom_line(aes(x = index, y = State, color = "State"), linewidth = 1) +
    labs(
      x = "index (weeks, 2022 Week 39 to 2025 Week 30)",
      y = "infected_value (?)",
      title = paste0("DTW-aligned Local vs. state inc values for HSA ID ", hsa_id, " from ", hsa$state)
    ) +
    theme_minimal() +
    scale_color_manual(name = "Legend", values = c("HSA (warped)" = "blue", State = "red"))
  
  (normal / warped) / (frechet + dtw)
}
```

```{r}
dtw_stats = function(hsa_id = 688) {
  hsa = hsa_state[hsa_state$hsa_nci_id == hsa_id, ]
  
  P = as.matrix(hsa$inc_hsa)
  Q = as.matrix(hsa$inc_state)
  
  res_dtw = dtw_distance(P, Q)
  
  res = discrete_frechet(P, Q)
  
  i = res_dtw$path[,1]
  j = res_dtw$path[,2]

  delta_t = j - i
  delta_x = P[i,1] - Q[j,1]

  stats_list = list(
    mean_abs_lag = mean(abs(delta_t)),
    signed_lag   = mean(delta_t),
    lag_sd       = sd(delta_t),
    dtw_mae      = mean(abs(delta_x)),
    dtw_rmse     = sqrt(mean(delta_x^2)),
    rmse = sqrt(mean((P[,1] - Q[,1])^2)),
    amp_ratio    = median(P[,1] / Q[,1], na.rm=TRUE),
    dtw_amp_ratio    = median(P[i,1] / Q[j,1], na.rm=TRUE),
    lag_amp_corr = cor(abs(delta_t), abs(delta_x)),
    frechet = res$frechet
  )

  stats = data.frame(
    Metric = names(stats_list),
    Value  = unname(unlist(stats_list))
  )
  
  return(stats)
}
```

```{r, fig.height = 14, fig.width = 10}
# set.seed(77)
id_list = hsa_state$hsa_nci_id
id = sample(id_list, 1)
graph_hsa(hsa_id = id)
knitr::kable(dtw_stats(hsa_id = id), digits = 3)  
```

```{r}
set.seed(77)

hsa_unique = hsa_state %>%
  distinct(hsa_nci_id, state)

state_sizes = hsa_unique %>%
  count(state, name = "N_state") %>%
  mutate(n_sample = ceiling(100 * N_state / length(hsa_unique$hsa_nci_id)))

hsa_list = list()
for (state_name in unique(hsa_state$state)) {
  state_subset = hsa_unique %>% filter(state == state_name)
  state_temp = sample(state_subset$hsa_nci_id, state_sizes[state_sizes$state == state_name, "n_sample"])
  hsa_list = append(hsa_list, state_temp)
}
```

```{r data, cache = TRUE}
res = vector("list", length(hsa_list))

for (i in seq_along(hsa_list)) {
  ID = hsa_list[[i]]
  
  out = pivot_wider(as.data.frame(dtw_stats(hsa_id = ID)), names_from  = Metric, values_from = Value)
  
  out$ID = ID
  out = out[, c("ID", setdiff(names(out), "ID"))]
  
  res[[i]] = out
}

dtw_results = do.call(rbind, res)
```

```{r}
dtw_results
knitr::kable(head(dtw_results))
```

Using our stratified sample, we can now look at some key statistics.
```{r}
metrics = setdiff(names(dtw_results), "ID")

summary = reshape(do.call(rbind, lapply(metrics, \(m) data.frame(Metric = m, Var2 = names(summary(dtw_results[[m]])), Freq = as.numeric(summary(dtw_results[[m]])), row.names = NULL))), idvar = "Metric", timevar = "Var2", direction = "wide")
colnames(summary) = c("Metric", "Minimum", "1st Quartile", "Median", "Mean", "3rd Quartile", "Maximum")
rownames(summary) = NULL
```

```{r, echo=FALSE}
var_explain = data.frame(
  Metric = c(
    "mean_abs_lag", "signed_lag", "lag_sd", "dtw_mae", "dtw_rmse", "rmse", "amp_ratio", "dtw_amp_ratio", "lag_amp_corr", "frechet"),
  Description = c(
    "Average absolute time lag between warped points",
    "Average time lag between warped points",
    "Standard deviation of time lag along DTW-warped path",
    "Mean absolute error along the DTW-warped path (shape comparison)",
    "Root mean squared error along the DTW-warped path (large differences penalized)",
    "Root mean squared error along the regular path",
    "Median ratio of amplitudes along the regular path (HSA % of state)",
    "Median ratio of amplitudes along the DTW-warped path (HSA % of state)",
    "Correlation between absolute time lag and amplitude warped difference",
    "Discrete Frechet distance (joint deviation in shape and timing)"
  ),
  Significance = c(
    "Larger values indicate greater overall timing mismatch between series",
    "Positive values imply HSA occurs later than State; negative values imply HSA leads State",
    "Larger values indicate more variable or unstable timing misalignment",
    "Larger values indicate greater amplitude mismatch after optimal alignment",
    "Penalizes large local mismatches, more sensitive to extreme deviations",
    "Baseline mismatch ignoring timing differences; comparison benchmark for DTW metrics",
     "Values > 1 indicate HSA generally larger than State; < 1 indicates smaller",
    "Values > 1 indicate HSA is larger than State along warping path; < 1 indicates smaller after alignment",
    "Positive values suggest larger timing shifts coincide with larger amplitude differences while negative values suggest that larger timing shifts coincide with smaller amplitude differences",
    "Lower values indicate more similar trajectories; higher values indicate greater overall deviation (leashed dog and owner, the Frechet distance is the smallest leash required)"
  ),
  stringsAsFactors = FALSE
)
```

```{r}
knitr::kable(as.data.frame(var_explain), caption = "Dynamic Time Warping Metric (+ Frechet) Descriptions and Significance") 
knitr::kable(summary %>% mutate(across(where(is.numeric), ~ round(., 3))), caption = "Summary Statistics for Dynamic Time Warping Metrics (+ Frechet) on HSA IDs sampled with stratification (state)") 
```

```{r, fig.height = 6, fig.width = 12}
dtw_results_long = dtw_results %>% pivot_longer(cols = -ID, names_to = "metric", values_to = "value")

ggplot(dtw_results_long, aes(x = value)) + geom_density(fill = "skyblue", alpha = 0.8) + facet_wrap(~ metric, scales = "free", nrow = 2) + 
  theme_minimal() + labs(title = "Density Plots for All DTW Metrics (+ Frechet)", x = "Value", y = "Density")
```

I think the most promising statistics are signed_lag, rmse - dtw_rmse (smaller is better), amp_ratio, and Frechet distance.
```{r, fig.height = 15, fig.width = 10}
hsa_dtw = dtw_results %>% left_join(hsa_state, by = join_by(ID == hsa_nci_id)) 
head(hsa_dtw)

hsa_dtw$ID = as.character(hsa_dtw$ID)

graph_lag = ggplot(hsa_dtw, aes(x = fct_reorder(state, lag_sd), y = lag_sd)) + 
  geom_boxplot(fill = "skyblue") + coord_flip() 

graph_rmse = ggplot(hsa_dtw, aes(x = fct_reorder(state, (rmse-dtw_rmse)), y = rmse-dtw_rmse)) + 
           geom_boxplot(fill = "skyblue") + coord_flip()
       
graph_amp = ggplot(hsa_dtw, aes(x = fct_reorder(state, amp_ratio), y = amp_ratio)) + 
          geom_boxplot(fill = "skyblue") + coord_flip()

graph_frechet = ggplot(hsa_dtw, aes(x = fct_reorder(state, frechet), y = frechet)) + 
           geom_boxplot(fill = "skyblue") + coord_flip()


(graph_lag + graph_rmse) / (graph_amp + graph_frechet)

ggplot(hsa_dtw, aes(x = fct_reorder(ID, lag_sd), y = lag_sd)) + 
  geom_point(fill = "skyblue") + coord_flip() 

```
Some cool box plots about some of the statistics, sorted. Might make some more after analyzing some of these metrics. I wanted to see any general trends in the data among states.
Refining the HSA dataset!
```{r}
hsa_dtw$density_ratio = hsa_dtw$density_hsa / hsa_dtw$density_state
hsa_dtw$area_ratio = hsa_dtw$area_km2_hsa / hsa_dtw$area_km2_state
colnames(hsa_dtw)

hsa_dtw_refined = hsa_dtw[,c(1:13, 16:17, 22, 29:30)]
```
This is a correlation matrix for the variables present in the DTW metrics and the socioeconomic variables.
```{r, fig.width = 8, fig.height = 8}
cor_matrix = cor(hsa_dtw_refined %>% select(where(is.numeric)))

corrplot(
  cor_matrix,
  method = "color",
  type = "lower",
  tl.col = "black",
  addCoef.col = "black",
    order = "hclust",
  number.cex = 0.6,
  diag = FALSE
)
```


# FluSight notes
During the epiENGAGE sessions in the first breakout session on Thursday, I was able to attend a session that focused on analyzing a few models from the FluSight dashboard. In particular, a few models that predicted exceptionally well were [CMU-TimeSeries](https://github.com/cmu-delphi/exploration-tooling/) and [Cornell_JHU-hierarchSIR](https://github.com/BentoLab-DiseaseDynamics/Cornell_JHU-hierarchSIR). 

## CMU-Time Series
The model is a AR model with specific data that CMU externally sources, I think. I found another link with more (https://zoltardata.com/model/307)

## Cornell_JHU-hierarchSIR
The model is a hybrid SIR - Bayesian hierarchical discrepancy model for infectious disease forecasting. After a quick run-through, I couldn't find much more about the specifics of their model but this is the type of model. 

































