---
title: "Report 5"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=4, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(patchwork)
library(zoo)
library(kableExtra)
library(corrplot)
```

# Introduction
This week, I will analyze the two FluSight models I have shortlisted and take some notes (which will be stored here). I will try to find some real trends (maybe with linear regression or a deeper analysis of that correlation plot) to look at DTW metrics in more detail. I also want to look at how trends happen over HSAs (? possibly because then I could take a look at some outliers) and look at how trends in states actually do different things. Also also, I think I can look at every row (HSA ID) now because the code is more optimized so that's nice. Looking forward to it!

# Data processing
```{r}
hsa_state = read.csv("../hsa_state_inc.csv")
```

## Metric functions
```{r}
source("../metric_functions.R")
source("../graphstat_functions.R")
source("../write_dtw.R")
```

```{r, fig.height = 14, fig.width = 10}
# set.seed(77)
# id_list = hsa_state$hsa_nci_id
# id = sample(id_list, 1)
# graph_hsa(hsa_id = id)
# knitr::kable(dtw_stats(hsa_id = id), digits = 3)  
graph_hsa()
```

```{r, cache = TRUE}
dtw_results = read_csv("../all_dtw_results.csv")
knitr::kable(head(dtw_results))
```

Using our sample, we can now look at some key statistics.
```{r}
summary = reshape(do.call(rbind, lapply(metrics, \(m) data.frame(Metric = m, Var2 = names(summary(dtw_results[[m]])), Freq = as.numeric(summary(dtw_results[[m]])), row.names = NULL))), idvar = "Metric", timevar = "Var2", direction = "wide")
colnames(summary) = c("Metric", "Minimum", "1st Quartile", "Median", "Mean", "3rd Quartile", "Maximum")
rownames(summary) = NULL

knitr::kable(summary %>% mutate(across(where(is.numeric), ~ round(., 3))), caption = "Summary Statistics for Dynamic Time Warping Metrics (+ Frechet) on HSA IDs sampled with stratification (state)") 
```

```{r, fig.height = 6, fig.width = 12}
dtw_results_long = dtw_results %>% pivot_longer(cols = -c(state, ID), names_to = "metric", values_to = "value")
hsa_unique = unique(hsa_state$state)

ggplot(dtw_results_long, aes(x = metric, y = value)) +
  geom_violin(fill = "royalblue", alpha = 0.6) +
  facet_grid(rows = vars(state), scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Metric Distributions Across States",
       x = "Metric", y = "Value")

```

```{r}
ggplot(dtw_results_long, aes(x = value)) + geom_density(fill = "skyblue", alpha = 0.8) + facet_wrap(~ metric, scales = "free", nrow = 2) + 
  theme_minimal() + labs(title = "Density Plots for All DTW Metrics (+ Frechet)", x = "Value", y = "Density")
```

# Violin plots
```{r, fig.length = 12, fig.width = 12}
id_state_lookup = hsa_state %>%
  select(hsa_nci_id, state) %>%
  distinct()

id_state_lookup$ID = id_state_lookup$hsa_nci_id

dtw_results_long = dtw_results_long %>%
  left_join(id_state_lookup, by = "ID")

ggplot(dtw_results_long, aes(x = metric, y = value)) +
  geom_violin(col = "") +
  facet_wrap(~state.x, ncol = 5, scales = "free_y")
```

# FluSight notes
During the epiENGAGE sessions in the first breakout session on Thursday, I was able to attend a session that focused on analyzing a few models from the FluSight dashboard. In particular, a few models that predicted exceptionally well were [CMU-TimeSeries](https://github.com/cmu-delphi/exploration-tooling/) and [Cornell_JHU-hierarchSIR](https://github.com/BentoLab-DiseaseDynamics/Cornell_JHU-hierarchSIR). 

## CMU-Time Series
The model is a AR model with specific data that CMU externally sources, I think. I found another link with more (https://zoltardata.com/model/307)

## Cornell_JHU-hierarchSIR
The model is a hybrid SIR - Bayesian hierarchical discrepancy model for infectious disease forecasting. After a quick run-through, I couldn't find much more about the specifics of their model but this is the type of model. 

































