---
title: "Report 4"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=4, fig.align = "center", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(patchwork)
library(corrplot)
library(mosaic)
library(moderndive)
library(effectsize)
library(tidyr)
library(caret)
library(purrr)
library(fastDummies)
library(GGally)
```

# Introduction
This week, I plan to investigate Bayesian additive regression trees in more detail through methods like hyperparamter tuning, a further deep dive into the literature surrounding the ML method, and improving data preprocessing and data collection. 

# Data preprocessing

# Implementation

# Variable importance with SHARP

# US map generation

# Literature review

# Conclusion