---
title: "Report 13"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=6, fig.width=7, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(stringr)
library(lubridate)
library(patchwork)
library(mosaic)
library(tidyr)
library(formatR)

library(data.table)
library(tidycensus)
library(dbarts)
library(iml)

library(xgboost)
library(pdp)
library(caret)
```

# Introduction
This week, I will be comparing the models against a baseline model, conducting residual analysis and displaying QQ-plots, creating PDP plots, testing the bias-variance error decomposition of both BART and XGBoost and lastly creating an ensemble of the two models and seeing how it performs relative to the baseline and the individual models. 


# Data preprocessing
```{r}
df_complete = read.csv("../complete_metrics.csv")
df_complete = drop_na(df_complete)

set.seed(123)
n = nrow(df_complete)
train_idx = sample(seq_len(n), size = floor(0.7*n)) # 80% train
test_idx = setdiff(seq_len(n), train_idx)

vars = c("pop_ratio", "pct_urban", "density_ratio", "n_hsa", "area_ratio", "education_diff", "income_diff", "poverty_diff", "unemployment_diff")

x.train = df_complete[train_idx, vars]
y.train = df_complete$diff_wis_season[train_idx]

x.test = df_complete[test_idx, vars]
y.test = df_complete$diff_wis_season[test_idx]
```

# Baseline model comparison
## Regular baseline
```{r, fig.width = 8, fig.height = 8}
baseline = lm(diff_wis_season ~ pop_ratio + pct_urban + density_ratio + n_hsa + area_ratio + education_diff + income_diff + poverty_diff + unemployment_diff, data = df_complete)
summary(baseline)

par(mfrow = c(2, 2))
plot(baseline)
par(mfrow = c(1, 1))
```

```{r}
pred_baseline = predict(baseline, newdata = df_complete)
obs = df_complete$diff_wis_season

R2 = 1 - sum((obs - pred_baseline)^2) / sum((obs - mean(obs))^2)
RMSE = sqrt(mean((obs - pred_baseline)^2))
```

```{r}
set.seed(123)

K = 10
folds = sample(rep(1:K, length.out = nrow(df_complete)))

cv_R2_values = c()

for (k in 1:K) {
  train_idx = which(folds != k)
  test_idx  = which(folds == k)

  m = lm(
    diff_wis_season ~ pop_ratio + pct_urban + density_ratio + n_hsa +
      area_ratio + education_diff + income_diff + poverty_diff + unemployment_diff,
    data = df_complete[train_idx, ]
  )

  pred_k = predict(m, newdata = df_complete[test_idx, ])
  obs_k  = df_complete$diff_wis_season[test_idx]

  R2_k = 1 - sum((obs_k - pred_k)^2) / sum((obs_k - mean(obs_k))^2)
  cv_R2_values[k] = R2_k
}

cv_mean_R2 = mean(cv_R2_values)
```

```{r, fig.height = 5}
baseline_predictions = data.frame(
  estimate = pred_baseline,
  observed = obs
)

ggplot(baseline_predictions, aes(x = estimate, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Predicted diff_wis_season",
    y = "Observed diff_wis_season",
    title = "Observed vs Predicted (season, baseline linear regression minus outlier)",
    subtitle = sprintf(
      "R-squared = %.2f | CV R-squared = %.2f | RMSE = %.2f | n = %.0f",
      R2, cv_mean_R2, RMSE, nrow(df_complete)
    )
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(-1, 2, by = 0.25)) +
  coord_cartesian(xlim = c(-0.25, 1))
```
## Baseline without outliers
```{r, fig.width = 8, fig.height = 8}
cooks = cooks.distance(baseline)
influential_idx = which(cooks > 4 / nrow(df_complete))
influential_idx

df_outlier = df_complete[-influential_idx, ]

baseline_outlier = lm(diff_wis_season ~ pop_ratio + pct_urban + density_ratio + n_hsa + area_ratio + education_diff + income_diff + poverty_diff + unemployment_diff, data = df_outlier)
summary(baseline_outlier)

par(mfrow = c(2, 2))
plot(baseline_outlier)
par(mfrow = c(1, 1))
```
```{r}
pred_baseline_outlier = predict(baseline_outlier, newdata = df_outlier)
obs_outlier = df_outlier$diff_wis_season

R2_outlier = 1 - sum((obs_outlier - pred_baseline_outlier)^2) / sum((obs_outlier - mean(obs_outlier))^2)
RMSE_outlier = sqrt(mean((obs_outlier - pred_baseline_outlier)^2))
```

```{r}
K = 10
folds = sample(rep(1:K, length.out = nrow(df_outlier)))

cv_R2_values = c()

for (k in 1:K) {
  train_idx = which(folds != k)
  test_idx  = which(folds == k)

  m = lm(
    diff_wis_season ~ pop_ratio + pct_urban + density_ratio + n_hsa +
      area_ratio + education_diff + income_diff + poverty_diff + unemployment_diff,
    data = df_outlier[train_idx, ]
  )

  pred_k = predict(m, newdata = df_outlier[test_idx, ])
  obs_k  = df_outlier$diff_wis_season[test_idx]

  R2_k = 1 - sum((obs_k - pred_k)^2) / sum((obs_k - mean(obs_k))^2)
  cv_R2_values[k] = R2_k
}

cv_mean_R2_outlier = mean(cv_R2_values)
```

```{r, fig.height = 5}
baseline_predictions_outlier = data.frame(
  estimate = pred_baseline_outlier,
  observed = obs_outlier
)

ggplot(baseline_predictions_outlier, aes(x = estimate, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Predicted diff_wis_season",
    y = "Observed diff_wis_season",
    title = "Observed vs Predicted (season, baseline linear regression)",
    subtitle = sprintf(
      "R-squared = %.2f | CV R-squared = %.2f | RMSE = %.2f | n = %.0f",
      R2_outlier, cv_mean_R2_outlier, RMSE, nrow(df_outlier)
    )
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(-1, 2, by = 0.25)) +
  coord_cartesian(xlim = c(-0.25, 1))
```

## Analysis
```{r, fig.height = 8, fig.width = 8}
pvals_baseline = summary(baseline)$coefficients[, "Pr(>|t|)"]
pvals_outlier  = summary(baseline_outlier)$coefficients[, "Pr(>|t|)"]

b = ggplot(data.frame(term = names(pvals_baseline), pvalue = pvals_baseline), aes(x = term, y = pvalue)
) + geom_col(fill = "skyblue", col="black") + labs(title = "Baseline Model Coefficient P-values", x = element_blank()) + theme_minimal() + theme(axis.text.x = element_blank()) 
  
bo = ggplot(data.frame(term = names(pvals_outlier), pvalue = pvals_outlier), aes(x = term, y = pvalue)
) + geom_col(fill = "tomato", col="black") + labs(title = "Outlier-Removed Baseline Model Coefficient P-values") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

b / bo
```
This baseline model mainly just serves to check how well the more complex models compare to a practically "naive" estimate of the difference between state and HSA. Both linear regression models (outlier and regular) both have n_hsa as the most important coefficient  followed by density_ratio and then education_diff. While not identical, they maintain the ordering for the top three most significant coefficients. However, the outlier-removed model drops density_ratio and education_diff p-values into a very significant range at below 0.01.

These baseline models and outlier-removal will be useful in the upcoming sections to test against the XGBoost and BART models since they have solid RMSE and $R^2$ metrics. For reference, the outlier-removed model improves in every metric.

# Partial dependence plots
In this section, we want to look at more feature importance related questions regarding our two most complex models. 

### BART
```{r, fig.width = 8, fig.height = 4}
bart_fit = bart(
  x.train = x.train,
  y.train = y.train,
  verbose = FALSE, keeptrees = TRUE)

pred_fun <- function(object, newdata) {
  preds <- predict(object, newdata = newdata)
  colMeans(preds)  
}
```

First, let's take a look at feature importance since it might be not the same as the baseline model. 
```{r}
bart_vi = as.data.frame(bart_fit$varcount) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "avg") 


ggplot(bart_vi) + geom_col(aes(x = reorder(variable, -avg), y = avg), fill='skyblue', col='black') + labs(x = "Variable", y = "Importance", title = "BART Variables by Importance") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Regular
```{r}
hsa.pdp = pdp::partial(object = bart_fit, pred.var = "n_hsa", pred.fun = pred_fun, train = df_complete, plot = FALSE) 
unemploy.pdp = pdp::partial(object = bart_fit, pred.var = "unemployment_diff", pred.fun = pred_fun, train = df_complete, plot = FALSE)  
```

```{r}
hsa = ggplot(hsa.pdp, aes(n_hsa, yhat)) + geom_line(aes(group = yhat.id), alpha = 0.2) +
stat_summary(fun = mean, geom = "line", col = "red", size = 1) + theme_minimal() + 
  labs(title = "ICE plot for n_hsa", x = "n_hsa", y = "Predicted response") + 
  scale_y_continuous(limits = c(-0.25,1.25), breaks = seq(-0.25, 1.25, by = 0.25))

unemploy = ggplot(unemploy.pdp, aes(unemployment_diff, yhat)) + geom_line(aes(group = yhat.id), alpha = 0.2) +
stat_summary(fun = mean, geom = "line", col = "red", size = 1) + theme_minimal() + 
  labs(title = "ICE plot for unemployment_diff", x = "unemployment_diff", y = "Predicted response") + 
  scale_y_continuous(limits = c(-0.25,1.25), breaks = seq(-0.25, 1.25, by = 0.25))
```

```{r, fig.width = 8, fig.height = 7}
hsa / unemploy
```

We can see that the different ICE lines have spread out along the "top" of these two figures while maintaining parallel to the PDP line and other ICE lines. That means that the effect might be changed drastically due to these outliers that maintain the global effect of these variables but skew the prediction.

#### Outliers
```{r}
n = nrow(df_outlier)
train_idx = sample(seq_len(n), size = floor(0.7*n)) # 80% train
test_idx = setdiff(seq_len(n), train_idx)

vars = c("pop_ratio", "pct_urban", "density_ratio", "n_hsa", "area_ratio", "education_diff", "income_diff", "poverty_diff", "unemployment_diff")

x.train2 = df_outlier[train_idx, vars]
y.train2 = df_outlier$diff_wis_season[train_idx]

x.test2 = df_outlier[test_idx, vars]
y.test2 = df_outlier$diff_wis_season[test_idx]

bart_fit2 = bart(
  x.train = x.train2,
  y.train = y.train2,
  verbose = FALSE, keeptrees = TRUE)
```

I also wanted to take a look at how removing the outliers would affect this since it makes education_diff and density_ratio much more significant in the linear baseline model.

```{r}
hsa.pdp2 = pdp::partial(object = bart_fit2, pred.var = "n_hsa", pred.fun = pred_fun, train = df_outlier, plot = FALSE) 
unemploy.pdp2 = pdp::partial(object = bart_fit2, pred.var = "unemployment_diff", pred.fun = pred_fun, train = df_outlier, plot = FALSE) 
```

```{r}
hsa2 = ggplot(hsa.pdp2, aes(n_hsa, yhat)) + geom_line(aes(group = yhat.id), alpha = 0.2) +
stat_summary(fun = mean, geom = "line", col = "red", size = 1) + theme_minimal() + 
  labs(title = "ICE plot for n_hsa (outlier-removed)", x = "n_hsa", y = "Predicted response") + 
  scale_y_continuous(limits = c(-0.25, 1), breaks = seq(-0.25, 1, by = 0.25))

unemploy2 = ggplot(unemploy.pdp2, aes(unemployment_diff, yhat)) + geom_line(aes(group = yhat.id), alpha = 0.2) +
stat_summary(fun = mean, geom = "line", col = "red", size = 1) + theme_minimal() + 
  labs(title = "ICE plot for unemployment_diff (outlier-removed)", x = "unemployment_diff", y = "Predicted response") + 
  scale_y_continuous(limits = c(-0.25, 1), breaks = seq(-0.25, 1, by = 0.25))
```

```{r, fig.width = 8, fig.height = 7}
hsa2 / unemploy2
```
For the top two metrics here, we can clearly see a steady decline for unemployment and a steady incline for n_hsa. Basically, the more HSAs there are, the larger the difference from HSA to state (that makes sense) and the larger the unemployment difference, the smaller the difference (this one is slightly confusing). However, we can see that removing the outliers enabled the ICE lines to coalesce more strongly around the center and the effect is global and more "together".


### XGBoost
```{r}
dtrain = xgb.DMatrix(data = as.matrix(x.train), label = y.train)
dtest  = xgb.DMatrix(data = as.matrix(x.test), label = y.test)

params = list(
  objective = "reg:squarederror",
  eval_metric = "rmse",
  eta = 0.05,              
  max_depth = 2,           
  subsample = 0.6,         
  colsample_bytree = 0.6,  
  min_child_weight = 5,    
  lambda = 5,              
  alpha = 2                
)

cv = xgb.cv(
  params = params,
  data = dtrain,
  nrounds = 500,
  nfold = 10,
  early_stopping_rounds = 20,
  verbose = 0
)

best_nrounds = cv$best_iteration

xgb_fit = xgb.train(
  params = params,
  data = dtrain,
  nrounds = best_nrounds,
  verbose = 0
)

pred_mean = predict(xgb_fit, as.matrix(x.test))

xgb_predictions = data.frame(
  estimate = pred_mean,
  observed = y.test
)

pred_fun_xgb <- function(object, newdata) {
  predict(object, as.matrix(newdata))
}
```

XGBoost has different important metrics. Let's check them out (gain is importance).
```{r}
knitr::kable(xgb.importance(model = xgb_fit))
```

```{r}
hsa.pdp.xgb = pdp::partial(object = xgb_fit, pred.var = "n_hsa", pred.fun = pred_fun_xgb, train = x.train, plot = FALSE) 
unemploy.pdp.xgb = pdp::partial(object = xgb_fit, pred.var = "unemployment_diff", pred.fun = pred_fun_xgb, train = x.train, plot = FALSE) 
```

```{r}
hsa.xgb = ggplot(hsa.pdp.xgb, aes(n_hsa, yhat)) + geom_line(aes(group = yhat.id), alpha = 0.2) +
stat_summary(fun = mean, geom = "line", col = "red", size = 1) + theme_minimal() + 
  labs(title = "ICE plot for n_hsa (XGBoost)", x = "n_hsa", y = "Predicted response") + 
  scale_y_continuous(limits = c(-0.25,1.25), breaks = seq(-0.25, 1.25, by = 0.25))

unemploy.xgb = ggplot(unemploy.pdp.xgb, aes(unemployment_diff, yhat)) + geom_line(aes(group = yhat.id), alpha = 0.2) +
stat_summary(fun = mean, geom = "line", col = "red", size = 1) + theme_minimal() + 
  labs(title = "ICE plot for unemployment_diff (XGBoost)", x = "unemployment_diff", y = "Predicted response") + 
  scale_y_continuous(limits = c(-0.25,1.25), breaks = seq(-0.25, 1.25, by = 0.25))
```

```{r, fig.width = 8, fig.height = 7}
hsa.xgb / unemploy.xgb
```
As we can see, the XGBoost model has a clear effect for n_hsa and unemployment_diff, however, these effects are global across all rows and therefore the PDP line represents the correlation quite well, with a dropoff or pickup at certain values of the variable that change the predicted response. Also, there are no very clear outliers in these figures like there was in BART, so no need to address that with the outlier-free versions.

# Why is $R^2$ so low?
```{r}
#y.train.scaled = scale(y.train)
#center = attr(y.train.scaled, "scaled:center")
#scale  = attr(y.train.scaled, "scaled:scale")

bart_fit = bart(
  x.train = x.train,
  y.train = y.train,
  verbose = FALSE, keeptrees = TRUE)

pred = predict(bart_fit, newdata = x.test)
bart_predictions = as.data.frame(t(apply(pred, 2, quantile, probs = c(0.025, 0.975))))
colnames(bart_predictions) = c("ci_low", "ci_high")

bart_predictions$estimate = apply(pred, 2, mean) 
#bart_predictions$estimate = bart_predictions$estimate * scale + center

bart_predictions$observed = y.test

var(bart_predictions$estimate) 
var(y.train)
var(y.test)

R2_bart = 1 - sum((bart_predictions$observed - bart_predictions$estimate)^2) /
  sum((bart_predictions$observed - mean(bart_predictions$observed))^2)
R2_bart
```
The variance of the training data is already very low and that makes sure that the $R^2$ values will be high since the variance is in the denominator. Also, the testing data variance is higher than the training data variance, which makes the predictive ability worse.

# Ensemble of both models
## Weighted average
Let's try a convex combination of weights and see what works the best!
```{r}
weights = seq(0, 1.5, by = 0.1)   # BART weights

results = data.frame(
  bart_weight = numeric(),
  xgb_weight = numeric(),
  R2 = numeric(),
  RMSE = numeric()
)

for (w1 in weights) {
  w2 = 1 - w1
  
  ensemble = w1 * bart_predictions$estimate + w2 * xgb_predictions$estimate
  
  R2_val = 1 - sum((y.test - ensemble)^2) /
               sum((y.test - mean(y.test))^2)
  
  RMSE_val = sqrt(mean((y.test - ensemble)^2))
  
  results = rbind(
    results,
    data.frame(
      bart_weight = w1,
      xgb_weight = w2,
      R2 = R2_val,
      RMSE = RMSE_val
    )
  )
}

knitr::kable(results, digits = 4, title = "Weighted average ensemble metrics")
```

Looks like a BART weight of 1.3 and a corresponding XGBoost weight of -0.3. Let's plot that real quick.
```{r}
w1 = 1.3 
w2 = -0.3 

ensemble_predictions = data.frame(
  estimate = w1 * bart_predictions$estimate + w2 * xgb_predictions$estimate,
  observed = y.test
)

R2_ensemble = 1 - sum((y.test - ensemble_predictions$estimate)^2) / sum((y.test - mean(y.test))^2)
RMSE_ensemble = sqrt(mean((y.test - ensemble_predictions$estimate)^2))
```

```{r, fig.width = 8, fig.height = 4}
ggplot(ensemble_predictions, aes(x = estimate, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Predicted diff_wis_season",
    y = "Observed diff_wis_season",
    title = "Observed vs Predicted (season, ensemble, 2)",
    subtitle = sprintf("R-squared = %.2f | RMSE = %.2f | n = %.0f", R2_ensemble, RMSE_ensemble, nrow(x.test))
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(-1, 2, by = 0.25)) +
  coord_cartesian(xlim = c(-0.25, 1))
```
Wow, this actually looks kinda decent and improves on the BART.

## Stacked
We can also try using linear regression to fit this for us.
```{r}
K = 5
folds = createFolds(y.train, k = K, returnTrain = FALSE)

bart_pred_train = rep(NA, length(y.train))
xgb_pred_train  = rep(NA, length(y.train))

for (i in 1:K) {
  test_idx  = folds[[i]]
  train_idx = setdiff(1:length(y.train), test_idx)
  
  x_tr = x.train[train_idx, ]
  y_tr = y.train[train_idx]
  
  x_te = x.train[test_idx, ]
  
  bart_fit_cv = bart(
    x_tr, y_tr,
    ntree = 200,
    verbose = FALSE,
    keeptrees = TRUE
  )
  bart_pred_train[test_idx] = colMeans(predict(bart_fit_cv, x_te))

  dtr = xgb.DMatrix(data = as.matrix(x_tr), label = y_tr)
  xgb_fit_cv = xgboost(
    data = dtr,
    nrounds = 200,
    verbose = 0
  )
  xgb_pred_train[test_idx] = predict(xgb_fit_cv, as.matrix(x_te))
}
```

```{r}
meta_train = data.frame(
  bart_pred = bart_pred_train,
  xgb_pred  = xgb_pred_train,
  y         = y.train
)

meta_model = lm(y ~ bart_pred + xgb_pred, data = meta_train)
coef(meta_model)

meta_test = data.frame(
  bart_pred = bart_predictions$estimate,
  xgb_pred  = xgb_predictions$estimate
)

ensemble_pred = predict(meta_model, newdata = meta_test)

ensemble_predictions_smart = data.frame(
  estimate = ensemble_pred,
  observed = y.test
)

R2_ensemble_smart = 1 - sum((y.test - ensemble_predictions_smart$estimate)^2) / sum((y.test - mean(y.test))^2)
RMSE_ensemble_smart = sqrt(mean((y.test - ensemble_predictions_smart$estimate)^2))
```

```{r, fig.width = 8, fig.height = 4}
ggplot(ensemble_predictions_smart, aes(x = estimate, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Predicted diff_wis_season",
    y = "Observed diff_wis_season",
    title = "Observed vs Predicted (season, ensemble, 2)",
    subtitle = sprintf("R-squared = %.2f | RMSE = %.2f | n = %.0f", R2_ensemble_smart, RMSE_ensemble_smart, nrow(x.test))
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(-1, 2, by = 0.25)) +
  coord_cartesian(xlim = c(-0.25, 1))
```

We can check how the ensemble is doing worse because both of these models are correlated with each other and therefore we are just amplifying noise right now.

```{r}
cor(bart_predictions$estimate, xgb_predictions$estimate)
```
Yikes, that's bad. Maybe next time or over break, I can investigate possibly trying to purposely overfit one of these models so that we can balance them out a bit. 

# Conclusion
We got to look at a baseline model that performed pretty well, a couple partial dependence plots and feature importance to analyze if there were any local discrepancies with the data, investigated why the $R^2$ was so low (answer: it was the very low variance in the test data combined with the overall lack of data), and then tried and kinda failed but kinda worked at an ensemble method combining BART and XGBoost. This was a nice semester and I think I'm proud of what we did. At the very least, BART gets a 75-80 percent coverage rate so that is very good from that perspective even if it can't explain all the variance in the testing data. We also got to explore a lot of interesting variables that turned out to have a significant effect, especially in the baseline, which was nice.