---
title: "Report 13"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=4, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(stringr)
library(lubridate)
library(patchwork)
library(mosaic)
library(tidyr)
library(formatR)

library(data.table)
library(tidycensus)
library(dbarts)
library(iml)

library(caret)
library(purrr)
library(fastDummies)
library(GGally)
library(mgcv) 

library(xgboost)
```

# Introduction
This week, I will be comparing the models against a baseline model, conducting residual analysis and displaying QQ-plots, creating PDP plots, testing the bias-variance error decomposition of both BART and XGBoost and lastly creating an ensemble of the two models and seeing how it performs relative to the baseline and the individual models. 


# Baseline model comparison

# Partial dependence plots

# Residual analysis
## QQ plots

# Bias-variance error decomposition
## BART
## XGBoost

# Ensemble of both models









