---
title: "Report 13"
author: "Advay Vyas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
urlcolor: blue
linkcolor: red
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=6, fig.width=7, fig.align = "left", warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

------------------------------------------------------------------------

```{r libraries, results='hide', warning=FALSE, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(stringr)
library(lubridate)
library(patchwork)
library(mosaic)
library(tidyr)
library(formatR)

library(data.table)
library(tidycensus)
library(dbarts)
library(iml)

library(xgboost)
library(pdp)
```

# Introduction
This week, I will be comparing the models against a baseline model, conducting residual analysis and displaying QQ-plots, creating PDP plots, testing the bias-variance error decomposition of both BART and XGBoost and lastly creating an ensemble of the two models and seeing how it performs relative to the baseline and the individual models. 


# Data preprocessing
```{r}
df_complete = read.csv("../complete_metrics.csv")
df_complete = drop_na(df_complete)

set.seed(123)
n = nrow(df_complete)
train_idx = sample(seq_len(n), size = floor(0.7*n)) # 80% train
test_idx = setdiff(seq_len(n), train_idx)

vars = c("pop_ratio", "pct_urban", "density_ratio", "n_hsa", "area_ratio", "education_diff", "income_diff", "poverty_diff", "unemployment_diff")

x.train = df_complete[train_idx, vars]
y.train = df_complete$diff_wis_season[train_idx]

x.test = df_complete[test_idx, vars]
y.test = df_complete$diff_wis_season[test_idx]
```

# Baseline model comparison
## Regular baseline
```{r, fig.width = 8, fig.height = 8}
baseline = lm(diff_wis_season ~ pop_ratio + pct_urban + density_ratio + n_hsa + area_ratio + education_diff + income_diff + poverty_diff + unemployment_diff, data = df_complete)
summary(baseline)

par(mfrow = c(2, 2))
plot(baseline)
par(mfrow = c(1, 1))
```

```{r}
pred_baseline = predict(baseline, newdata = df_complete)
obs = df_complete$diff_wis_season

R2 = 1 - sum((obs - pred_baseline)^2) / sum((obs - mean(obs))^2)
RMSE = sqrt(mean((obs - pred_baseline)^2))
```

```{r}
set.seed(123)

K = 10
folds = sample(rep(1:K, length.out = nrow(df_complete)))

cv_R2_values = c()

for (k in 1:K) {
  train_idx = which(folds != k)
  test_idx  = which(folds == k)

  m = lm(
    diff_wis_season ~ pop_ratio + pct_urban + density_ratio + n_hsa +
      area_ratio + education_diff + income_diff + poverty_diff + unemployment_diff,
    data = df_complete[train_idx, ]
  )

  pred_k = predict(m, newdata = df_complete[test_idx, ])
  obs_k  = df_complete$diff_wis_season[test_idx]

  R2_k = 1 - sum((obs_k - pred_k)^2) / sum((obs_k - mean(obs_k))^2)
  cv_R2_values[k] = R2_k
}

cv_mean_R2 = mean(cv_R2_values)
```

```{r, fig.height = 5}
baseline_predictions = data.frame(
  estimate = pred_baseline,
  observed = obs
)

ggplot(baseline_predictions, aes(x = estimate, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Predicted diff_wis_season",
    y = "Observed diff_wis_season",
    title = "Observed vs Predicted (season, baseline linear regression)",
    subtitle = sprintf(
      "R-squared = %.2f | CV R-squared = %.2f | RMSE = %.2f | n = %.0f",
      R2, cv_mean_R2, RMSE, nrow(df_complete)
    )
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(-1, 2, by = 0.25)) +
  coord_cartesian(xlim = c(-0.25, 1))
```
## Baseline without outliers
```{r, fig.width = 8, fig.height = 8}
cooks = cooks.distance(baseline)
influential_idx = which(cooks > 4 / nrow(df_complete))
influential_idx

df_outlier = df_complete[-influential_idx, ]

baseline_outlier = lm(diff_wis_season ~ pop_ratio + pct_urban + density_ratio + n_hsa + area_ratio + education_diff + income_diff + poverty_diff + unemployment_diff, data = df_outlier)
summary(baseline_outlier)

par(mfrow = c(2, 2))
plot(baseline_outlier)
par(mfrow = c(1, 1))
```
```{r}
pred_baseline_outlier = predict(baseline_outlier, newdata = df_outlier)
obs_outlier = df_outlier$diff_wis_season

R2_outlier = 1 - sum((obs_outlier - pred_baseline_outlier)^2) / sum((obs_outlier - mean(obs_outlier))^2)
RMSE_outlier = sqrt(mean((obs_outlier - pred_baseline_outlier)^2))
```

```{r}
K = 10
folds = sample(rep(1:K, length.out = nrow(df_outlier)))

cv_R2_values = c()

for (k in 1:K) {
  train_idx = which(folds != k)
  test_idx  = which(folds == k)

  m = lm(
    diff_wis_season ~ pop_ratio + pct_urban + density_ratio + n_hsa +
      area_ratio + education_diff + income_diff + poverty_diff + unemployment_diff,
    data = df_outlier[train_idx, ]
  )

  pred_k = predict(m, newdata = df_outlier[test_idx, ])
  obs_k  = df_outlier$diff_wis_season[test_idx]

  R2_k = 1 - sum((obs_k - pred_k)^2) / sum((obs_k - mean(obs_k))^2)
  cv_R2_values[k] = R2_k
}

cv_mean_R2_outlier = mean(cv_R2_values)
```

```{r, fig.height = 5}
baseline_predictions_outlier = data.frame(
  estimate = pred_baseline_outlier,
  observed = obs_outlier
)

ggplot(baseline_predictions_outlier, aes(x = estimate, y = observed)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Predicted diff_wis_season",
    y = "Observed diff_wis_season",
    title = "Observed vs Predicted (season, baseline linear regression)",
    subtitle = sprintf(
      "R-squared = %.2f | CV R-squared = %.2f | RMSE = %.2f | n = %.0f",
      R2_outlier, cv_mean_R2_outlier, RMSE, nrow(df_outlier)
    )
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(-1, 2, by = 0.25)) +
  coord_cartesian(xlim = c(-0.25, 1))
```

## Analysis
```{r, fig.height = 8, fig.width = 8}
pvals_baseline = summary(baseline)$coefficients[, "Pr(>|t|)"]
pvals_outlier  = summary(baseline_outlier)$coefficients[, "Pr(>|t|)"]

b = ggplot(data.frame(term = names(pvals_baseline), pvalue = pvals_baseline), aes(x = term, y = pvalue)
) + geom_col(fill = "skyblue", col="black") + labs(title = "Baseline Model Coefficient P-values", x = element_blank()) + theme_minimal() + theme(axis.text.x = element_blank()) 
  
bo = ggplot(data.frame(term = names(pvals_outlier), pvalue = pvals_outlier), aes(x = term, y = pvalue)
) + geom_col(fill = "tomato", col="black") + labs(title = "Outlier-Removed Baseline Model Coefficient P-values") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

b / bo
```
This baseline model mainly just serves to check how well the more complex models compare to a practically "naive" estimate of the difference between state and HSA. Both linear regression models (outlier and regular) both have n_hsa as the most important coefficient  followed by density_ratio and then education_diff. While not identical, they maintain the ordering for the top three most significant coefficients. However, the outlier-removed model drops density_ratio and education_diff p-values into a very significant range at below 0.01.

These baseline models and outlier-removal will be useful in the upcoming sections to test against the XGBoost and BART models since they have solid RMSE and $R^2$ metrics. For reference, the outlier-removed model improves in every metric.

# Partial dependence plots

# Residual analysis
## QQ plots

# Bias-variance error decomposition
## BART
## XGBoost

# Ensemble of both models









